---
title: "Validating a new bmm model"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
library(bmmtools)
res <- readRDS(from_root("vignettes/articles/assets/bmmtools.rds"))
```

This article walks once through the loop every bmmtools check is built
on: simulate data from known parameters, fit the model, score the fit
against the truth. It uses one model and one design, so that each step
is visible. [Running a recovery study](recovery-grid.html) repeats the
same loop over a design grid.

The second half asks the three other questions a new model has to
answer. `prior_check()` asks what the priors imply before any data are
fitted, `sbc()` whether the implementation is correct, and
`cross_check()` whether the model agrees with what is already known.
Only the scorer changes; the loop is the same one.

The model is bmm's two-parameter mixture model, `mixture2p`, because an
adapter for it ships with bmmtools. A model without one works the same
way once you supply a generator; see
[Validating a model without a built-in adapter](own-model.html).

To run the code you need bmm, brms and a working
[cmdstanr](https://mc-stan.org/cmdstanr/) installation. The output below
was computed once with that code and saved, so the page itself was built
without Stan.

## The model and the truth

```{r model, eval = FALSE, code = shown("model", "vignettes/articles/precompute/bmmtools.R")}
```

`pars` are the population values that generate the data. They are given
**on the link scale** and under bmm's own parameter names, because that
is the scale the model is estimated on: `kappa` has a log link and
`thetat` a logit link, so `log(8)` is a precision of 8 and
`qlogis(0.75)` a mixture weight of 0.75. `sds` are between-subject
standard deviations on the same scale. A parameter not named in `sds`
does not vary between subjects.

## Simulate

```{r simulate, eval = FALSE, code = shown("simulate", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
res$data
res$truth$population
```

`simulate_recovery()` draws one value per subject and parameter on the
link scale, converts it with `inverse_link()` and hands it to the
model's own generator, here `bmm::rmixture2p()`. It returns the data
and the truth: `truth$population` holds the values in `pars`, and
`truth$subjects` holds each subject's drawn values, both on the link
scale. The seed makes the draw reproducible and is applied only around
it, so the global random number state is left alone.

## Fit once, reuse afterwards

```{r fit, eval = FALSE, code = shown("fit", "vignettes/articles/precompute/bmmtools.R")}
```

`recovery_formula()` gives every free parameter of the model a random
intercept over `id`, the column `simulate_recovery()` writes.
`fit_cached()` fits with `bmm::bmm()` and saves the fit next to a key.
Calling it again with the same arguments reads the fit back instead of
sampling:

```{r refit, eval = FALSE, code = shown("refit", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
res$reused
```

The key covers the formula, data, model, prior, seed, chains,
iterations, warmup, thinning, `control`, `init`, the backend and the
installed versions of bmm, brms and the Stan toolchain. When one of
them changes, the model is refitted and the message names what changed.

## The convergence gate

```{r convergence, eval = FALSE, code = shown("convergence", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
res$convergence
```

`check_convergence()` returns the worst rhat, the smallest bulk and tail
effective sample sizes, the divergent transitions and the tree-depth
hits over all variables, and a verdict under thresholds that are
arguments: rhat at most 1.05, bulk ESS at least 400 and at most ten
divergent transitions by default. A fit that fails is still scored, and
the verdict travels with the estimates as the `converged` column.

## Score the population

```{r recover, eval = FALSE, code = shown("recover", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
res$population
```

`recover()` extracts the posterior median and a 95% equal-tailed
credible interval per parameter, joins the truth, and scores. Scoring
is on the natural scale by default: the estimate, its interval and the
truth are converted with the links the `bmmfit` records, so `kappa` is
a precision and `thetat` a probability.

With a single fit there is one estimate per parameter, so bias and RMSE
are that one error, coverage is 0 or 1, and the correlations are `NA`
because a correlation needs at least three pairs. Population-level
recovery needs many fits, which is what `recovery_grid()` produces.

## Score the subjects

```{r subjects, eval = FALSE, code = shown("subjects", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
summary(res$subjects)
```

`recover_subjects()` does the same for each subject's estimate, which
is the population intercept plus the subject's deviation, summed per
posterior draw. One fit now gives 40 pairs per parameter, so the
correlations are defined. `summary()` returns one row per parameter;
[Recovery summary columns](metrics.html) defines every column.

## Plot

```{r plot, eval = FALSE, code = shown("plot", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
plot_recovery(res$subjects)
```

The generating value is on the x axis, the posterior median on the y
axis, each bar is a credible interval and the line marks equality.
Panels are per parameter with free scales.

## Checking the priors first

A recovery study fits the model's priors many times, so it is worth
seeing what they imply for the data before the first fit.
`prior_check()` samples from the prior only and summarises the
prior-predictive draws on the scale of the response:

```{r, eval = FALSE}
checked <- prior_check(model, recovery_formula(model), sim$data, seed = 1)
summary(checked)
plot_prior_check(checked)
```

[Prior predictive checks on the observable scale](prior-check.html)
covers the summaries, custom statistics and comparing prior sets.

## Is the implementation correct?

Recovery asks whether the model can be estimated from data of a given
size. It cannot tell you whether the likelihood you wrote is the one you
meant: a model with a sign error can still recover its own parameters.
Simulation-based calibration (SBC) asks the other question. If the
implementation is correct, then over data sets drawn from the prior the
rank of the true value among the posterior draws is uniform, and a
histogram of those ranks is flat.

`sbc()` fits the prior once, turns each prior draw into a data set
through `simulate_recovery()`, fits every one of them, and hands the
result to the [SBC](https://hyunjimoon.github.io/SBC/) package, which
owns the ranks and the plots. It needs a data layout rather than data:
the responses are generated, so only the design matters.

```{r sbc-prior, eval = FALSE, code = shown("sbc-prior", "vignettes/articles/precompute/bmmtools.R")}
```

The prior is given explicitly, and that is not incidental. bmm's default
prior on the between-subject SD of `kappa` is a half-`student_t(3, 0,
2.5)` on the **log** scale. Over twenty draws it reached 9.50, which puts
a subject two SDs above the mean at a concentration of about `exp(21)`,
and `bmm::rmixture2p()` fails on it. Two of twenty prior draws could not
be generated from at all. A prior that puts mass where the model's own
generator breaks cannot be used for SBC, and that is a finding about the
prior rather than about the model: SBC draws from the prior, so a prior
wide enough to be meaningless on the observable scale shows up here
first.

```{r sbc, eval = FALSE, code = shown("sbc", "vignettes/articles/precompute/bmmtools.R")}
```

`level` picks which draws are ranked, here the two population parameters
and the two between-subject SDs. Whatever is ranked, everything the
formula implies is always drawn from the prior and simulated from,
because the ranks are uniform only when the data come from the joint
prior.

```{r sbc-plot, eval = FALSE, code = shown("sbc-plot", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE, fig.height = 4}
SBC::plot_rank_hist(res$ranks)
```

The run ranked `r length(res$sbc_info$variables)` variables over
`r res$sbc_info$diagnostics$n_fits` simulations, with
`r res$sbc_info$diagnostics$n_high_rhat` fits above bmmtools' rhat bar of
`r res$sbc_info$diagnostics$rhat_max` and
`r res$sbc_info$diagnostics$n_low_ess_to_rank` below SBC's rank-ESS bar,
and no backend errors.

**Twenty simulations is a smoke test, not a verdict.** It shows that the
pipeline runs end to end, that the truths and the draws are named
alike, and that nothing is grossly wrong. It cannot tell a flat
histogram from a mildly sloped one: with 20 draws across
`r unique(res$ranks$stats$max_rank)[[1]] + 1` rank bins, almost any shape
is consistent with uniformity. Answering whether `mixture2p` is
calibrated needs a few hundred simulations and hours of sampling. Run
`sbc()` at `n_sims = 20` while you are still changing the model, and
once at a serious size before you believe the answer.

`sbc()` returns SBC's own `SBC_results`, so every SBC function applies
unchanged: `SBC::plot_ecdf_diff()`, `SBC::plot_sim_estimated()` and the
rest.

## Does it agree with what is known?

The last question is whether a new model agrees with what the field
already has. `cross_check()` compares a fit's estimates with a
reference: a closed form, another implementation, or values from a
published paper.

`mixture2p` has no closed form to compare against, so this section uses
signal detection instead, where one has existed since the 1960s. The
model is `sdt_yn`, fitted to simulated yes-no data:

```{r cross-check-fit, eval = FALSE, code = shown("cross-check-fit", "vignettes/articles/precompute/bmmtools.R")}
```

The reference is the textbook estimator computed from the observed hit
and false-alarm rates, here
`r sprintf("%.3f", res$hit_rate)` and `r sprintf("%.3f", res$fa_rate)`:

```{r cross-check, eval = FALSE, code = shown("cross-check", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
res$checked
```

Printing summarises, as it does for the other scorers: `bias`, `rmse`
and `coverage` are `summary.bmmtools_recovery()`'s words for the same
quantities, so the three scorers read with one vocabulary. With one fit
there is a single comparison per parameter, so `rmse` is the absolute
bias and `coverage` is 0 or 1.

The object itself is one row per parameter, and holds the comparison
term by term:

```{r, eval = FALSE}
checked[, c(
  "term", "estimate", "ci_low", "ci_high",
  "reference", "bias", "covered", "overlap"
)]
```

```{r, echo = FALSE}
res$checked[, c(
  "term", "estimate", "ci_low", "ci_high",
  "reference", "bias", "covered", "overlap"
)]
```

`bias` is the signed difference from the reference and `covered` says
whether the fit's credible interval contains it. `overlap` is `NA`
because a closed form has no interval of its own; with a reference that
does have one — published values with a confidence interval, say — it
reports whether the two intervals meet, and it is never `FALSE` for a
reference without one, so a missing interval cannot read as a
disagreement.

Selecting columns dropped the `bmmtools_cross_check` class, which is
deliberate: the class promises the full contract, and a table that no
longer carries it is a plain tibble.

The wording matters more than it looks. **The reference is a comparison,
not a truth.** The closed form makes its own assumptions, and a
disagreement means the two disagree, not that the model is wrong. Here
the equal-variance assumption holds by construction, because `sdt_yn`
fixes `sdratio`, so `d` is d′ and the two are directly comparable. With
`sdratio` free the model estimates d~a~ instead, and the reference has
to be converted before it is comparable; `?cross_check` gives the
conversion.

```{r cross-check-plot, eval = FALSE, code = shown("cross-check-plot", "vignettes/articles/precompute/bmmtools.R")}
```

```{r, echo = FALSE}
plot_recovery(res$checked)
```

The reference is on the x axis, the fit's credible interval is vertical,
and a reference with its own interval would draw a horizontal one.

## Which function does what

| Step | Function | Returns |
| --- | --- | --- |
| simulate | `simulate_recovery()` | data and truth |
| formula | `recovery_formula()` | a `bmmformula` with random intercepts |
| fit | `fit_cached()` | the fit, reused while nothing changed |
| gate | `check_convergence()` | one row of diagnostics and a verdict |
| score | `recover()`, `recover_subjects()` | a `bmmtools_recovery` tibble |
| summarise | `summary()` | one row per parameter (and condition) |
| plot | `plot_recovery()`, `plot_prior_check()` | a ggplot |
| loop | `recovery_grid()` | all of the above over a design grid |
| priors | `prior_check()` | prior-predictive statistics |
| calibration | `sbc()` | SBC's `SBC_results`, ranks and their plots |
| agreement | `cross_check()` | a `bmmtools_cross_check` tibble |
