---
title: "Recovering between-subject correlations"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
script <- "vignettes/articles/precompute/correlation-recovery.R"
library(bmmtools)
res <- readRDS(from_root("vignettes/articles/assets/correlation-recovery.rds"))
cors <- res$cor_recovery

# numbers in the text: two decimals, no leading zero for values within [-1, 1]
num <- function(x, digits = 2) {
  out <- formatC(x, digits = digits, format = "f")
  ifelse(abs(x) < 1, sub("^(-?)0\\.", "\\1.", out), out)
}
# one cell of by_type
cell <- function(type, r, cor, est, column) {
  row <- by_type[by_type$pair_type == type & by_type$rho == r &
    by_type$re_cor == cor & by_type$estimator == est, ]
  num(row[[column]])
}
lav <- function(r, cor, pair, column) {
  s <- res$lavaan_grid_summary
  num(s[[column]][s$rho == r & s$re_cor == cor & s$pair == pair])
}
lav_status <- function(cond, status) {
  s <- res$lavaan_status
  n <- s$n[s$condition == cond & s$status == status]
  if (length(n) == 0L) 0L else n
}
groups <- dplyr::left_join(res$convergence_groups, res$design,
  by = "condition"
)
group_fits <- function(cor, grp, column) {
  g <- groups[groups$re_cor == cor & groups$group == grp, ]
  sum(g[[column]] > 0)
}
att <- function(term, est, column) {
  a <- res$attenuation
  num(a[[column]][a$term == term & a$estimator == est])
}
```

A model that measures individual differences has to do more than order
people correctly on each of its parameters. Most substantive questions
are about how those differences relate: whether precision in one task
goes with precision in another, whether two parameters of the same model
are separable, or whether a parameter correlates with an ability
measured outside the model. [Validating a new bmm model](bmmtools.html)
scored how well subject estimates recover the subjects' true values.
This article scores how well the correlations between them are
recovered.

> **A first look, not a benchmark.** The results below come from one
> small simulation: four conditions, 20 replications each, 100 subjects
> per data set, fitted with four chains of 1,000 iterations, half of them
> warmup. At that length
> many fits miss the usual convergence standard, and we score them under
> a looser one (see [Convergence](#convergence)). The results show how
> the tools work and in which direction each estimator errs. They are
> not a calibrated recommendation for study design. A more comprehensive
> simulation will replace this article once the package is stable.

To run the code you need bmm, brms, a working
[cmdstanr](https://mc-stan.org/cmdstanr/) installation and, for the SEM
part, lavaan. The output was computed once and saved, so this page was
built without Stan or lavaan. Simulating and fitting the grid took
`r round(res$minutes_fitting)` minutes.

## Generating correlated people

```{r model, eval = FALSE, code = shown("model", script)}
```

The model is bmm's two-parameter mixture model, with population values
and between-subject SDs on the link scale as in
[Validating a new bmm model](bmmtools.html). Each simulated subject
does two tasks with 100 trials each, so there are four subject-level
terms, `kappa` and `thetat` in task 1 and task 2, and one observed
covariate `G`, standing in for an ability measured by a separate test.

The correlations among these five variables come from a factor model:

```{r truth, code = shown("truth", script)}
```

`cors_from_factors()` turns loadings and factor correlations into the
implied correlation matrix. The two tasks load $\sqrt{.7}$ on a factor
per parameter, `precision` for `kappa` and `memory` for `thetat`, and
`G` loads 1 on its own factor `ability`. All three factors correlate
$\rho$. This gives three kinds of pairs with known values:

- **the same parameter across tasks**, $.7$ at every $\rho$;
- **different parameters**, $.7\rho$, so $.35$ at $\rho = .5$;
- **a parameter with `G`**, $\sqrt{.7}\,\rho$, so $.42$ at $\rho = .5$.

We used a factor model, rather than one correlation for every pair,
for two reasons. First, the article can then separate the correlation
between two terms, which the estimators below score, from the
correlation between the latent factors, which a structural equation
model recovers. Second, the factor model keeps that SEM identified even
at $\rho = 0$, because the two tasks of one parameter still correlate
$.7$.

## Three estimators

`extract_correlations()` offers three answers to the question of how two
subject-level terms correlate:

- `"model"` is the correlation of the random effects the model
  estimates. It exists only when the random effects of the two terms
  share a correlation block, `(1 | p | id)`.
- `"draws"` correlates the subjects' values within each posterior draw
  and summarises these correlations like any other posterior, with the
  median and a 95% equal-tailed credible interval (CrI).
- `"point"` correlates the subjects' posterior means, with a Fisher-z
  interval that treats the means as if they were observed values.

They target different quantities. `model` estimates the correlation in
the population the subjects come from. `draws` describes the subjects in
the data set: its interval reflects uncertainty about their values, not
about which subjects happened to be sampled. `point` is a correlation of
estimates, and so it inherits whatever shrinkage and estimation error
those estimates carry. `recover_correlations()` therefore scores each
estimate against two truths: `true_value`, the generating correlation,
and `sample_value`, the correlation the simulated subjects actually had.

## The grid

```{r grid, eval = FALSE, code = shown("grid", script)}
```

The grid crosses the factor correlation, $\rho = 0$ or $.5$, with the
random-effects structure of the fitted model. Under
`re_cor = "none"` every term has its own random intercept and the model
assumes the terms are uncorrelated. Under `re_cor = "all"` the four
terms share one correlation block, so the model estimates their
correlations.

```{r run, eval = FALSE, code = shown("run", script)}
```

`cors` and `formula` are functions of the grid row. `cors_for()` builds
the correlation matrix for the row's $\rho$, and the formula function
reads the row's `re_cor`, which is how one grid compares two model
specifications. `tasks` and `covariates` add the task dimension and `G`
to the simulated data. `correlations` asks the grid to extract all three
estimators from every fit; `model` rows exist only in the
`re_cor = "all"` cells. `threads` enables within-chain parallelisation
in brms; in a timing test with one 100-subject fit it cut the sampling
time from 499 to 238 seconds.

## Convergence {#convergence}

```{r convergence, eval = FALSE, code = shown("convergence", script)}
```

```{r, echo = FALSE}
convergence_table <- res$convergence |>
  dplyr::summarise(
    fits = dplyr::n(),
    pass = sum(pass),
    pass_strict = sum(pass_strict),
    max_rhat = max(max_rhat),
    min_ess_bulk = min(min_ess_bulk),
    min_ess_tail = min(min_ess_tail),
    n_divergent = sum(n_divergent),
    .by = condition
  ) |>
  dplyr::left_join(res$design, by = "condition")
convergence_table[c(
  "condition", "rho", "re_cor", "fits", "pass",
  "pass_strict", "max_rhat", "min_ess_bulk",
  "min_ess_tail", "n_divergent"
)]
```

Under the usual standard, Rhat below 1.01, bulk and tail ESS above 400
and no divergent transitions, only `r sum(res$convergence$pass_strict)`
of the `r nrow(res$convergence)` fits pass. None had a divergent
transition or hit the maximum tree depth; they fall short on Rhat and
ESS, and not by much. The worst Rhat is
`r num(max(res$convergence$max_rhat))`, the smallest bulk ESS
`r round(min(res$convergence$min_ess_bulk))` and the smallest tail ESS
`r round(min(res$convergence$min_ess_tail))`. We therefore score every
fit with Rhat below `r res$bar$rhat_max` and bulk and tail ESS above
`r res$bar$ess_bulk_min`, which all `r sum(res$convergence$pass)` fits
meet.

The shortfall is concentrated in a few variables:

```{r convergence-groups, eval = FALSE, code = shown("convergence-groups", script)}
```

```{r, echo = FALSE}
groups |>
  dplyr::summarise(
    fits = dplyr::n(),
    fits_rhat = sum(n_rhat > 0),
    fits_ess = sum(n_ess > 0),
    max_rhat = max(max_rhat),
    min_ess_bulk = min(min_ess_bulk),
    .by = c(re_cor, group)
  )
```

In the `re_cor = "all"` cells, `r group_fits("all", "cor", "n_rhat")` of
40 fits have a correlation parameter with Rhat of 1.01 or above, and
`r group_fits("all", "cor", "n_ess")` have one with ESS at or below 400.
The smallest bulk ESS of any correlation parameter is
`r round(min(groups$min_ess_bulk[groups$group == "cor"]))`. These are
the draws the `model` estimator summarises, so its intervals below rest
on fewer effective draws than the other estimators'. In the
`re_cor = "none"` cells, `r group_fits("none", "subject", "n_rhat")` of
40 fits have at least one of their 400 subject-level terms with Rhat of
1.01 or above, `r group_fits("none", "sd", "n_rhat")` an SD and
`r group_fits("none", "population", "n_rhat")` a population parameter.
Longer chains are the obvious remedy. For a simulation meant to show
the direction of errors rather than exact calibration, we did not run
them.

## Scoring

```{r correlations, eval = FALSE, code = shown("correlations", script)}
```

`summary()` of the correlation recovery has one row per condition, pair
and estimator. With ten pairs, three estimators and four conditions this
is a long table, so we average it within the three kinds of pairs:

```{r}
pair_type <- function(term) {
  dplyr::case_when(
    grepl("(^|__)G(__|$)", term) ~ "with G",
    grepl("^kappa.*__kappa|^thetat.*__thetat", term) ~ "same parameter",
    TRUE ~ "across parameters"
  )
}
by_type <- res$cor_summary |>
  dplyr::mutate(pair_type = pair_type(term)) |>
  dplyr::summarise(
    true_value = mean(true_value),
    mean_estimate = mean(mean_estimate),
    bias = mean(bias),
    coverage = mean(coverage),
    coverage_sample = mean(coverage_sample),
    rejection_rate = mean(rejection_rate),
    ci_width = mean(ci_width),
    .by = c(pair_type, rho, re_cor, estimator)
  )
print(by_type, n = Inf)
```

`rejection_rate` is the share of replications whose interval excludes
zero: the false-positive rate where the true correlation is 0 and the
power where it is not. [Recovery summary columns](metrics.html) defines
the other columns.

```{r, fig.height = 4.5}
library(ggplot2)
ggplot(
  by_type,
  aes(estimator, mean_estimate, colour = re_cor)
) +
  geom_hline(aes(yintercept = true_value), linetype = "dashed") +
  geom_point(size = 2.5, position = position_dodge(width = 0.4)) +
  facet_grid(paste("rho =", rho) ~ pair_type) +
  labs(
    x = NULL, y = "Mean estimate (dashed: generating value)",
    colour = "re_cor"
  )
```

**The same parameter across tasks.** The generating correlation is .70.
A model without correlated random effects shrinks each subject towards
the mean of that term alone, and both estimators are attenuated: `draws`
gives `r cell("same parameter", 0.5, "none", "draws", "mean_estimate")`
and `point` `r cell("same parameter", 0.5, "none", "point", "mean_estimate")`
at $\rho = .5$. With correlated random effects, `model` and `draws`
recover the correlation, at
`r cell("same parameter", 0.5, "all", "model", "mean_estimate")` and
`r cell("same parameter", 0.5, "all", "draws", "mean_estimate")`.
`point` now overshoots, at
`r cell("same parameter", 0.5, "all", "point", "mean_estimate")`, and
because its interval is narrow (mean width
`r cell("same parameter", 0.5, "all", "point", "ci_width")`) it covers
the generating value in only
`r cell("same parameter", 0.5, "all", "point", "coverage")` of the
replications. A likely reason is that the correlated model shrinks
each subject towards the correlation it estimates, so the posterior
means line up more closely than the true values do.

**Different parameters.** At $\rho = 0$ this pair type measures false
positives. Under `re_cor = "all"`, `point` excludes zero in
`r cell("across parameters", 0, "all", "point", "rejection_rate")` of
the replications, against
`r cell("across parameters", 0, "all", "draws", "rejection_rate")` for
`draws` and
`r cell("across parameters", 0, "all", "model", "rejection_rate")` for
`model`. At $\rho = .5$ (generating value .35) the same pattern holds:
`draws` and `model` are close,
`r cell("across parameters", 0.5, "all", "draws", "mean_estimate")` and
`r cell("across parameters", 0.5, "all", "model", "mean_estimate")`,
and `point` is inflated to
`r cell("across parameters", 0.5, "all", "point", "mean_estimate")`.
Under `re_cor = "none"`, `draws` is attenuated to
`r cell("across parameters", 0.5, "none", "draws", "mean_estimate")`,
while `point` lands near the truth at
`r cell("across parameters", 0.5, "none", "point", "mean_estimate")`.
However, the same `point` estimator gives
`r cell("across parameters", 0, "none", "point", "mean_estimate")`
where the truth is 0, so this is not an unbiased estimator but, in this
design, an attenuation and an upward bias that partly cancel.

**A parameter with `G`.** Here `point` is the best calibrated
estimator. At $\rho = 0$ it excludes zero in
`r cell("with G", 0, "all", "point", "rejection_rate")` of the
replications under `re_cor = "all"`, and at $\rho = .5$ its mean is
`r cell("with G", 0.5, "all", "point", "mean_estimate")` against a
generating value of .42, covered in
`r cell("with G", 0.5, "all", "point", "coverage")` of the replications.
`draws` excludes zero in
`r cell("with G", 0, "all", "draws", "rejection_rate")` of the
replications at $\rho = 0$. This follows from its target: its
interval covers the correlation these 100 subjects actually had in
`r cell("with G", 0, "all", "draws", "coverage_sample")` of the
replications. A sample of 100 subjects has a correlation with `G` that
differs from 0 by chance, and the `draws` interval, which carries no
uncertainty about which subjects were sampled, is narrow enough
(mean width `r cell("with G", 0, "all", "draws", "ci_width")`) to
detect it.

Taken together, in this design, correlations between parameters of one
model were best recovered by fitting correlated random effects and
reading `model` or `draws`. Correlating posterior means from that model
overstated the correlations and inflated false positives. For a
correlation with an observed covariate, `point` answered the population
question and `draws` the in-sample one.

`plot_recovery()` shows the replications of one cell, each estimate
against the correlation the subjects actually had:

```{r plot, fig.width = 9, fig.height = 7.5, code = shown("plot", script)}
```

## The SEM route

A second way to estimate a latent correlation is to take the subject
estimates out of the model and fit a structural equation model to them.
`subject_table()` returns one row per subject with the true and
estimated value of every term and the covariates:

```{r subject-table, eval = FALSE, code = shown("subject-table", script)}
```

```{r lavaan-grid, eval = FALSE, code = shown("lavaan-grid", script)}
```

The CFA mirrors the generating model. Each parameter's two tasks load
on one factor with equal loadings, `G` is observed, and within a task
the residuals of `kappa` and `thetat` covary, because both are estimated
from the same trials. Every fit is kept with its status:

```{r, echo = FALSE}
res$lavaan_status
lavaan_summary <- res$lavaan_grid_summary[
  c(
    "condition", "rho", "re_cor", "pair", "n_fits", "n_warning",
    "mean_estimate", "sd_estimate", "coverage"
  )
]
dplyr::mutate(lavaan_summary, dplyr::across(mean_estimate:coverage, \(x) round(x, 3)))
```

Under `re_cor = "none"` all 40 CFAs fit without warnings. The
correlations with `G` are recovered, at
`r lav(0.5, "none", "precision__G", "mean_estimate")` and
`r lav(0.5, "none", "memory__G", "mean_estimate")` for a latent .5, and
`r lav(0, "none", "precision__G", "mean_estimate")` and
`r lav(0, "none", "memory__G", "mean_estimate")` for 0. The correlation
between the two parameter factors is not: it is
`r lav(0, "none", "precision__memory", "mean_estimate")` where the
factors are uncorrelated and
`r lav(0.5, "none", "precision__memory", "mean_estimate")` where they
correlate .5. This is consistent with the upward bias of `point`
between parameters seen above, which the CFA then corrects for
attenuation and so enlarges.

Under `re_cor = "all"`, `r lav_status("row-3", "warning")` of 20 CFAs at
$\rho = 0$ and `r lav_status("row-4", "warning")` of 20 at $\rho = .5$
end with a warning: negative residual variances or a residual covariance
matrix that is not positive definite. The factor correlation varies
widely between replications, with an SD of
`r lav(0.5, "all", "precision__memory", "sd_estimate")` at $\rho = .5$,
and its 95% confidence interval covers .5 in
`r lav(0.5, "all", "precision__memory", "coverage")` of them. One explanation, which this simulation does not test, is
that posterior means from a model with correlated random effects
already carry the correlation the model estimated, and a CFA on them
has little room left to model it again.

In this design, the two-step route recovers correlations with an
observed covariate but not the correlation between two parameters of
the same model. lavaan ran in the precompute script; it is not a
dependency of bmmtools.

## Several models, separate fits

Sometimes the terms come from different models, fitted separately:
precision in a visual working memory task and in a verbal one, each
with its own `bmm` fit. `recovery_component()` describes one model,
`simulate_components()` draws correlated subjects across several, and
`fit_components()` fits each:

```{r components, eval = FALSE, code = shown("components", script)}
```

```{r fit-components, eval = FALSE, code = shown("fit-components", script)}
```

```{r, echo = FALSE}
res$set_convergence
```

This example is a single data set of 100 subjects at $\rho = .5$, so
its numbers illustrate the arithmetic rather than estimate a bias. Both
fits pass the looser bar; fit `a` misses the strict one on Rhat.

```{r separate-fits, eval = FALSE, code = shown("separate-fits", script)}
```

```{r reliability, eval = FALSE, code = shown("reliability", script)}
```

```{r, echo = FALSE}
res$attenuation[c(
  "term", "estimator", "true_value", "mean_estimate",
  "rel_a", "rel_b", "expected", "disattenuated"
)]
```

Neither fit knows about the other model's subjects, so a correlation
across fits is attenuated by the reliabilities of the two subject
estimates. Here the reliability is the squared correlation of true and
estimated values, between
`r num(min(res$reliability))` and `r num(max(res$reliability))`. For
`kappa` the generating correlation of .70 comes out as
`r att("a_kappa__b_kappa", "point", "mean_estimate")` with `point` and
`r att("a_kappa__b_kappa", "draws", "mean_estimate")` with `draws`,
close to the expected
`r att("a_kappa__b_kappa", "point", "expected")` for
$\rho\sqrt{rel_a\,rel_b}$ and
`r att("a_kappa__b_kappa", "draws", "expected")` for
$\rho\,rel_a\,rel_b$. Dividing by those factors gives
`r att("a_kappa__b_kappa", "point", "disattenuated")` and
`r att("a_kappa__b_kappa", "draws", "disattenuated")`. For `thetat`
both estimates lie above the expected values,
`r att("a_thetat__b_thetat", "point", "mean_estimate")` against
`r att("a_thetat__b_thetat", "point", "expected")` and
`r att("a_thetat__b_thetat", "draws", "mean_estimate")` against
`r att("a_thetat__b_thetat", "draws", "expected")`, and their
disattenuated values,
`r att("a_thetat__b_thetat", "point", "disattenuated")` and
`r att("a_thetat__b_thetat", "draws", "disattenuated")`, overshoot .70.
A single data set cannot tell whether this is noise. In practice the
reliabilities are not known and would have to be estimated as well.

```{r lavaan-components, eval = FALSE, code = shown("lavaan-components", script)}
```

```{r, echo = FALSE}
dplyr::mutate(
  res$lavaan_set[c("pair", "estimate", "ci_low", "ci_high", "status")],
  dplyr::across(estimate:ci_high, \(x) round(x, 3))
)
```

The CFA on the separate fits recovers the correlations with `G`,
`r num(res$lavaan_set$estimate[res$lavaan_set$pair == "precision__G"])`
and
`r num(res$lavaan_set$estimate[res$lavaan_set$pair == "memory__G"])`,
and overestimates the correlation of the two parameter factors at
`r num(res$lavaan_set$estimate[res$lavaan_set$pair == "precision__memory"])`,
the same direction as in the grid. A joint multivariate fit of both
models would estimate these correlations directly; bmmtools does not
support that yet.

## Which function does what

| Step | Function | Returns |
| --- | --- | --- |
| truth | `cors_from_factors()` | a correlation matrix implied by a factor model |
| simulate | `simulate_recovery()`, `simulate_components()` | data and truth, with tasks and covariates |
| formula | `recovery_formula(re_cor = )` | random intercepts, correlated or not |
| fit | `fit_cached()`, `fit_components()` | fits, reused while nothing changed |
| gate | `check_convergence()` | diagnostics and a verdict under your thresholds |
| extract | `extract_correlations()` | `model`, `draws` and `point` correlations |
| score | `recover_correlations()` | estimates against true and sample correlations |
| summarise | `summary()` | one row per pair, estimator and condition |
| two-step | `subject_table()` | true and estimated values per subject, for an SEM |
| loop | `recovery_grid(correlations = )` | all of the above over a design grid |
