---
title: "Hierarchical estimation against subject-wise maximum likelihood"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
library(bmmtools)
res <- readRDS(from_root("vignettes/articles/assets/hierarchical-vs-ml.rds"))
fmt <- function(x, digits = 3) sprintf(paste0("%.", digits, "f"), x)
# every number in the prose comes from one table cell, and a selection
# that does not match exactly one row is an error rather than the first
# match: a change to the precompute must break the render, not the text
one <- function(tab, keys, column) {
  row <- Reduce(`&`, Map(function(k, v) tab[[k]] == v, names(keys), keys))
  if (sum(row) != 1L) {
    stop(sprintf(
      "%s of %s matched %d rows, not 1",
      column, paste(names(keys), unlist(keys), sep = " = ", collapse = ", "),
      sum(row)
    ))
  }
  tab[[column]][row]
}
pick <- function(tab, term, estimator, column) {
  one(tab, list(term = term, estimator = estimator), column)
}
cell <- function(term, estimator, n_trials, column) {
  one(
    res$by_trials,
    list(term = term, estimator = estimator, n_trials = n_trials),
    column
  )
}
link_summary <- summary(res$scored_link)
nat <- summary(res$scored)
```

A hierarchical model estimates each subject partly from that subject's
own data and partly from everyone else's. That this buys accuracy is
easy to assert and rarely measured on the model in front of you. This
article measures it for one bmm model: the same simulated data fitted
twice, once hierarchically and once subject by subject with no pooling,
and scored once against the generating values that produced it.

`fit_ml()` is the second fit. It exists for this comparison and not for
inference: bmm's own discussion of subject-wise maximum likelihood
records the wish that users not apply it for inference, and bmmtools
does not recommend it either. What it is good for is a baseline, and a
baseline is what turns "hierarchical estimation helps" into a number.

The comparison is the one
[Recovery summary columns](metrics.html#reading-the-concordance) works
through algebraically for a conjugate normal model. Here the same
quantities come out of a real bmm model, a real sampler and the
package's own scorers.

## The data

The data set is the one [Validating a new bmm model](bmmtools.html)
walks through: `mixture2p`, 40 subjects, 100 trials each, subjects
drawn around a population value on the link scale.

```{r simulate, eval = FALSE, code = shown("simulate", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

## Two fits of it

The hierarchical fit gives every free parameter a random intercept over
`id`, which is what `recovery_formula()` writes:

```{r hierarchical, eval = FALSE, code = shown("hierarchical", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$convergence
```

`fit_ml()` takes the hierarchy out. It writes `p ~ 0 + id` instead, so
each subject gets an independent set of parameters and nothing is
shared between them:

```{r ml, eval = FALSE, code = shown("ml", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$ml
```

Two things about that fit are what keep it a fair opponent rather than a
straw man.

**The priors are flat.** Under bmm's default priors the mode of a
no-pooling fit is a penalised estimate, and the comparison would become
hierarchical shrinkage against prior shrinkage, which is a different and
much weaker claim. `prior = "default"` is available and documented as
what it is.

**It is one fit, not forty.** Under `p ~ 0 + id` every free parameter
belongs to a single subject, so the log posterior is a sum of
per-subject terms and the joint mode is the vector of per-subject
modes. `?fit_ml` reports the check: the joint mode and independent
optimisation of bmm's own density agreed to 3.0e-06 on the link scale.

## One truth, two estimators

There is no `compare_estimators()`. Both fits produce an estimates
tibble with the same columns, the rows carry an `estimator` column, and
binding them is the join:

```{r bind, eval = FALSE, code = shown("bind", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$bound
```

`recover_subjects()` then scores all `r nrow(res$scored_link)` rows
against the one set of generating values, and `summary()` groups by
`estimator` instead of pooling the two into a single bias and RMSE:

```{r score-link, eval = FALSE, code = shown("score-link", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
link_summary[c(
  "term", "estimator", "n", "bias", "rmse", "coverage",
  "ci_width", "r"
)]
link_summary[c(
  "term", "estimator", "r", "ccc", "ccc_scale_shift",
  "calibration_slope", "truth_sd"
)]
```

That is one `summary()` object, printed twice over different columns;
all 23 are defined in [Recovery summary columns](metrics.html).

`scale = "link"` is deliberate here. The link scale is the one the model
is estimated on and the one the shrinkage argument is about; the natural
scale comes back below, and it does not tell the same story.

## What the table says

**RMSE falls and the correlation does not.** Subject-level RMSE goes
from `r fmt(pick(link_summary, "kappa", "ml", "rmse"))` to
`r fmt(pick(link_summary, "kappa", "bayes", "rmse"))` on `kappa` and
from `r fmt(pick(link_summary, "thetat", "ml", "rmse"))` to
`r fmt(pick(link_summary, "thetat", "bayes", "rmse"))` on `thetat`,
while `r` moves from `r fmt(pick(link_summary, "kappa", "ml", "r"))` to
`r fmt(pick(link_summary, "kappa", "bayes", "r"))` and from
`r fmt(pick(link_summary, "thetat", "ml", "r"))` to
`r fmt(pick(link_summary, "thetat", "bayes", "r"))`. This is the shape
the conjugate table in [Recovery summary columns](metrics.html) shows
with exact posterior means: the same ordering of subjects, recovered
with less error.

**They differ far more in spread than in rank.** `ccc_scale_shift`
is the standard deviation of the estimates over the standard deviation
of the truth. Printed rather than inferred:

```{r spread, eval = FALSE, code = shown("spread", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$spread
```

The maximum-likelihood estimates spread wider than the subjects they
estimate, because each one carries its own measurement error and nothing
pulls it back. The hierarchical estimates spread narrower, because each
is pulled towards the population mean by an amount the model infers from
how noisy that subject's data are.

**`ccc` does not settle it, and that is a property of `ccc`.** On
`kappa` Lin's concordance is
`r fmt(pick(link_summary, "kappa", "ml", "ccc"))` for maximum likelihood
against `r fmt(pick(link_summary, "kappa", "bayes", "ccc"))` for the
hierarchical fit, the wrong way round from the RMSE. Lin's
bias-correction factor treats a scale shift of $v$ and one of $1/v$
alike, so $\rho_c$ cannot distinguish calibrated shrinkage from none and
mildly rewards too little of it. [Recovery summary
columns](metrics.html#reading-the-concordance) derives this for the
conjugate case, where it shows up as .691 against .657; here it survives
contact with a real model. Read `ccc` next to `rmse`, never instead of
it.

**One replication is a coarse instrument for `calibration_slope`.** For
a calibrated posterior mean the slope of truth on estimate is 1.
Measured here it is `r fmt(pick(link_summary, "kappa", "bayes", "calibration_slope"))`
on `kappa` and `r fmt(pick(link_summary, "thetat", "bayes", "calibration_slope"))`
on `thetat`, against
`r fmt(pick(link_summary, "kappa", "ml", "calibration_slope"))` and
`r fmt(pick(link_summary, "thetat", "ml", "calibration_slope"))` for
maximum likelihood. The direction is unambiguous and the distance from 1
is not: the slope is $r / v$, and both are estimated from
`r pick(link_summary, "kappa", "bayes", "n")` subjects in a single
simulated data set. The design grid below is where that number becomes
readable.

A warning that applies to the whole table. `ccc_scale_shift` near 1 is
**not** the target for the hierarchical estimator. A calibrated
posterior mean has $v = r$, not $v = 1$; `calibration_slope` $= r/v$ is
the column to read against 1. This is documented in `?recovery_ccc` and
derived in [Recovery summary
columns](metrics.html#reading-the-concordance).

```{r plot, eval = FALSE, code = shown("plot", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE, fig.height = 4}
plot_recovery(res$scored_link, color_by = "estimator", annotate = TRUE)
```

Both clouds sit on the identity line. The maximum-likelihood cloud is
longer along it and its intervals are wider.

## On the scale a reader interprets

The link scale is the honest one for comparing estimators. It is not the
one results are reported on. `kappa` is a concentration and `thetat` a
probability, and the hierarchical fit records its own link table while a
`fit_ml()` result does not, so scoring a bound tibble on the natural
scale needs the links supplied:

```{r links, eval = FALSE, code = shown("links", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$links
```

```{r score-natural, eval = FALSE, code = shown("score-natural", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
nat[c(
  "term", "estimator", "n", "bias", "rmse",
  "coverage", "r", "calibration_slope"
)]
```

`bias` on `kappa` is where the two scales part company. On the link
scale both estimators are near-unbiased,
`r fmt(pick(link_summary, "kappa", "ml", "bias"), 4)` and
`r fmt(pick(link_summary, "kappa", "bayes", "bias"), 4)`. On the natural
scale they are biased in opposite directions,
`r fmt(pick(nat, "kappa", "ml", "bias"))` for maximum likelihood and
`r fmt(pick(nat, "kappa", "bayes", "bias"))` for the hierarchical fit.

Nothing has gone wrong; this is Jensen's inequality and the spread
column above. `kappa` has a log link and `exp()` is convex, so the mean
of the transformed estimates depends on their spread as well as their
centre. The unpooled estimates spread `r fmt(pick(res$spread, "kappa", "ml", "ratio"), 2)`
times as wide as the truth and come out too high; the shrunk ones spread
`r fmt(pick(res$spread, "kappa", "bayes", "ratio"), 2)` times as wide
and come out too low. RMSE still prefers the hierarchical fit on this
scale, `r fmt(pick(nat, "kappa", "bayes", "rmse"), 2)` against
`r fmt(pick(nat, "kappa", "ml", "rmse"), 2)`. Report the scale next to
the number, and read bias on the scale the model was estimated on.

## What maximum likelihood cannot be scored on

The hierarchical fit also estimates the population, and there is nothing
in the no-pooling fit to compare it against:

```{r population, eval = FALSE, code = shown("population", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$population
```

With one fit there is one estimate per parameter, so `rmse` is the
absolute bias, `coverage` is 0 or 1, and the correlations are `NA`.
Averaging the subject-wise estimates would produce a number, but not one
with an interval that accounts for how well each subject was measured.
That is a second thing hierarchical estimation buys, and it is not in
the table above because it cannot be.

## The same answer without a compiler

`method = "optim"` maximises bmm's R density for the model directly and
takes its interval from the Hessian. It needs no CmdStan and no
compilation:

```{r optim, eval = FALSE, code = shown("optim", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$optim_summary[c(
  "term", "estimator", "n", "bias", "rmse",
  "coverage", "r", "calibration_slope"
)]
```

The maximum-likelihood rows are the same rows to the Monte Carlo error
of the Laplace draws:
`r fmt(pick(res$optim_summary, "kappa", "ml", "rmse"))` against
`r fmt(pick(link_summary, "kappa", "ml", "rmse"))` on RMSE and
`r fmt(pick(res$optim_summary, "kappa", "ml", "calibration_slope"))`
against
`r fmt(pick(link_summary, "kappa", "ml", "calibration_slope"))` on the
calibration slope, both for `kappa`. `coverage` moves by one subject in
each parameter. Row by row, exactly two of the 80 rows change their
verdict, and in both the generating value lies within 0.01 of an
interval bound: the Laplace interval is quantiles of 1000 draws and the
Wald interval is analytic, so the two do not agree to the last digit and
a borderline subject can fall either side. The route is a cost, not a moderator of the
result: `?fit_ml` records the check that the Wald and Laplace standard
errors agree to a mean ratio of 1.000 over the same 40 subjects. It is capped at the models bmmtools carries a density for, the
same seven the simulation adapters cover, and `nll` supplies one for
anything else.

## Over a design

One data set gives one number per cell of the table. `recovery_grid()`
runs the whole comparison over a design, fitting both estimators on the
same simulated data in every cell, and `ml = list(...)` passes arguments
to `fit_ml()`. The design below varies the quantity the shrinkage
argument is about, how much data each subject contributes:

```{r grid, eval = FALSE, code = shown("grid", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

`ml = list(method = "optim")` means the whole maximum-likelihood side of
this grid needs no compiler. Of the
`r fmt(res$grid_secs + res$grid_ml_secs, 0)` seconds the cells took,
`r fmt(res$grid_ml_secs, 1)` were the ML fits.

The grid scores on the natural scale by default. Running it again with
`scale = "link"` re-reads the stored per-cell estimates and refits
nothing:

```{r grid-link, eval = FALSE, code = shown("grid-link", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r grid-summary, eval = FALSE, code = shown("grid-summary", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$by_trials[c(
  "term", "estimator", "n_trials", "n", "n_converged",
  "bias", "rmse", "r", "coverage", "calibration_slope"
)]
```

Three things in that table are the point of running a grid at all.

**The gap is a function of how much each subject contributes.** On
`kappa`, RMSE runs `r fmt(cell("kappa", "ml", 25, "rmse"))` against
`r fmt(cell("kappa", "bayes", 25, "rmse"))` at 25 trials and
`r fmt(cell("kappa", "ml", 100, "rmse"))` against
`r fmt(cell("kappa", "bayes", 100, "rmse"))` at 100. Pooling buys most
where each subject is measured worst, which is what it is for, and the
two estimators converge as the per-subject data grow.

**The calibration slope now reads.** For maximum likelihood it climbs
from `r fmt(cell("kappa", "ml", 25, "calibration_slope"))` at 25 trials
to `r fmt(cell("kappa", "ml", 100, "calibration_slope"))` at 100, all of
them far below 1: noisy estimates spread wider than the subjects they
describe, and less so as the noise falls. The hierarchical slopes sit
either side of 1 with no trend. Neither statement could be made from the
single fit above.

**All `r nrow(res$cells)` hierarchical fits converged; the ML fits did
not.** The `n` column is
`r cell("kappa", "ml", 25, "n")` subjects per replication for maximum
likelihood at 25 trials against `r cell("kappa", "bayes", 25, "n")` for
the hierarchical fit.

```{r grid-plot, eval = FALSE, code = shown("grid-plot", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE, fig.height = 3.5}
library(ggplot2)
ggplot(res$by_trials, aes(factor(n_trials), rmse, colour = estimator)) +
  geom_point(size = 2.5) +
  geom_line(aes(group = estimator)) +
  facet_wrap(~term, scales = "free_y") +
  labs(
    x = "Trials per subject", y = "Subject-level RMSE (link scale)",
    colour = "Estimator"
  )
```

## When a subject's fit fails

A subject whose estimate leaves a finite range on the link scale, or
whose optimiser reports failure, keeps its row with `estimate = NA` and
`converged = FALSE`. It is not dropped, and the reason matters: the
recovery metrics drop incomplete pairs silently, so dropping the failures
would compare the hierarchical fit on every subject against maximum
likelihood on the subjects it found easy, and report the difference as
the estimator's.

What that leaves is a comparison on unequal `n`, which is why `n` sits
in the summary next to `bias` and `rmse` and why the per-cell counts are
kept:

```{r ml-cells, eval = FALSE, code = shown("ml-cells", "vignettes/articles/precompute/hierarchical-vs-ml.R")}
```

```{r, echo = FALSE}
res$ml_cells
```

`attr(x, "cells")` is the hierarchical side of the same table, with the
seed, file, runtime and convergence verdict per cell.

Which way the missing subjects push matters. A subject whose maximum
likelihood estimate runs to a boundary is a subject whose data were
least informative, and a metric that drops it is computed on the
subjects maximum likelihood found easiest while the hierarchical row
beside it uses all
`r cell("kappa", "bayes", 25, "n")`. The 25-trial comparison above
therefore **understates** the gap rather than inventing it. There is no
setting in which dropping them would have been the safer choice, which
is why `fit_ml()` has none.

## What this is not

**It is not a recommendation.** Nothing here says to fit models subject
by subject. It says what you give up if you do, on this model, at these
sample sizes.

**It is not a general result.** The size of the gap depends on the
model, on how much data each subject provides, on how much subjects
really differ, and on whether the generating values come from the
population the hierarchical model assumes. Here they do, by
construction, which is the most favourable case for shrinkage.
`recovery_grid()` is how you find out what happens for your own model
and design.

**A single replication settles less than it looks.** With 40 subjects,
`r` and the scale shift are both noisy, so `calibration_slope` is noisy
twice over. Read it from a grid with replications, not from one fit.

**Coverage is not the whole of calibration.** A Laplace or Wald interval
around a maximum-likelihood estimate can cover the generating value at
close to its nominal rate while the point estimate is far worse, and in
this data set it does. Coverage says the interval is honest about the
estimate's own uncertainty, not that the estimate is good.
