---
title: "Recovery summary columns"
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
library(bmmtools)
```

This page defines every column `summary()` returns for a
`bmmtools_recovery` object: what is computed, over which rows, and when
the value is `NA`. It describes the computation only.

```{r}
summary(recovery_mixture2p)[1:4, ]
```

## Which rows each row summarises

One summary row covers one parameter (`term`) at one `level`, and one
`condition` when the object comes from `recovery_grid()`. Write
$\hat\theta_i$ for the estimate in row $i$ (the posterior median),
$\theta_i$ for the generating value, and $[l_i, u_i]$ for the credible
interval, all on the scale in the `scale` column.

At the **population** level the rows are the replications, one estimate
per fit. At the **subject** level the rows are subjects within
replications; the error metrics use all of them, and the correlation
metrics are computed within each replication and then combined, as
described below.

Missing values are dropped pairwise before anything is counted, so `n`
always reports the pairs a metric used.

## Identification columns

`term`, `level`, `scale`
: The parameter, `"population"` or `"subject"`, and `"natural"` or
  `"link"`.

`n`
: Complete estimate–truth pairs. At subject level, the mean number of
  subjects per replication.

`n_replications`
: Distinct replications contributing to the row.

`n_converged`
: Replications whose fit passed `check_convergence()` with the
  thresholds in force when the estimates were extracted. `NA`, not 0,
  when no row carries a verdict, as with a hand-built estimates tibble.

## Error metrics

`bias`
: The mean signed error, $\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)$.

`rmse`
: The root mean squared error,
  $\sqrt{\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)^2}$.

`coverage`
: The share of intervals containing the generating value,
  $\frac{1}{n}\sum_i \mathbb{1}[l_i \le \theta_i \le u_i]$. The interval
  is closed, so a bound equal to the truth counts as covered. Its
  nominal value is the `ci_level` of the estimates, 0.95 by default.

`ci_width`
: The mean interval width, $\frac{1}{n}\sum_i (u_i - l_i)$.

These four are `NA` only when no complete pair exists.

## Correlation metrics

`r`, `r_low`, `r_high`
: The Pearson correlation between estimates and truth with a 95%
  confidence interval from Fisher's $z$ transformation,
  $z = \operatorname{atanh}(r)$ with standard error $1/\sqrt{n-3}$,
  transformed back with $\tanh$. The interval is always 95% and does not
  follow `ci_level`, which is the mass of the posterior intervals.

`rank_r`
: The Spearman rank correlation.

`ccc`
: Lin's concordance correlation coefficient [@linConcordanceCorrelationCoefficient1989],
  computed with population moments (dividing by $n$):
  $$\rho_c = \frac{2 s_{\hat\theta\theta}}{s^2_{\hat\theta} + s^2_\theta + (\bar{\hat\theta} - \bar\theta)^2}.$$

`ccc_scale_bias`
: The ratio of standard deviations, $s_{\hat\theta} / s_\theta$.

`ccc_location_bias`
: The mean difference relative to the geometric mean of the two
  standard deviations,
  $(\bar{\hat\theta} - \bar\theta) / \sqrt{s_{\hat\theta}\, s_\theta}$.

All correlation metrics are `NA`, never 0, when fewer than three
complete pairs exist or when estimates or truth have no spread. This is
why population-level correlations are `NA` in a grid whose replications
share their population values. With exactly three pairs, `r` has a point
estimate and no interval.

### Subject level: within replication, then combined

At subject level, each replication is correlated on its own. The
Pearson and Spearman correlations are then combined on Fisher's $z$
scale with weights $n_k - 3$, where $n_k$ is the number of subjects in
replication $k$:

$$\bar z = \frac{\sum_k (n_k - 3)\, z_k}{\sum_k (n_k - 3)}, \qquad \operatorname{SE}(\bar z) = \frac{1}{\sqrt{\sum_k (n_k - 3)}}.$$

`r` is $\tanh(\bar z)$ and its interval is $\tanh(\bar z \pm 1.96\,
\operatorname{SE})$. With one replication this is the ordinary Fisher
interval. `rank_r` is combined the same way, without an interval.
`ccc` and its two components are averaged across replications as they
are, since they are not correlations on the $z$ scale.

## The scale

`recover()` and `recover_subjects()` score on the natural scale by
default: estimates, interval bounds and truth are converted with
`inverse_link()` before any metric is computed, and `scale = "link"`
skips the conversion. Coverage does not depend on the choice, because
a monotone transformation keeps a value inside its interval, but bias,
RMSE, interval width and the correlations do.

Two links need care at zero. Under the `sqrt` link, an interval that
spans zero on the link scale becomes $[0, \max(l^2, u^2)]$. Under the
`inverse` link, an interval that spans zero has no image on the natural
scale, so its bounds and `covered` are `NA` and a warning names the
term.

## Column names

The first nine columns of a recovery object (`term`, `estimate`,
`ci_low`, `ci_high`, `ci_method`, `ci_level`, `rhat`, `ess_bulk`,
`ess_tail`) follow the column contract of the
[apabayes](https://github.com/GidonFrischkorn/apabayes) package, so
recovery tables can be formatted there without renaming.

## References
