---
title: "Recovery summary columns"
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
library(bmmtools)
```

This page defines every column `summary()` returns for a
`bmmtools_recovery` object: what is computed, over which rows, and when
the value is `NA`. It describes the computation only.

```{r}
s <- summary(recovery_mixture2p)
s[1:4, 1:14]
s[1:4, c(1:3, 15:23)]
```

## Which rows each row summarises

One summary row covers one parameter (`term`) at one `level`, and one
`condition` when the object comes from `recovery_grid()`. Write
$\hat\theta_i$ for the estimate in row $i$ (the posterior median),
$\theta_i$ for the generating value, and $[l_i, u_i]$ for the credible
interval, all on the scale in the `scale` column.

At the **population** level the rows are the replications, one estimate
per fit. At the **subject** level the rows are subjects within
replications; the error metrics use all of them, and the correlation
metrics are computed within each replication and then combined, as
described below.

Missing values are dropped pairwise before anything is counted, so `n`
always reports the pairs a metric used.

## Identification columns

`term`, `level`, `scale`
: The parameter, `"population"` or `"subject"`, and `"natural"` or
  `"link"`.

`n`
: Complete estimate–truth pairs. At subject level, the mean number of
  subjects per replication.

`n_replications`
: Distinct replications contributing to the row.

`n_converged`
: Replications whose fit passed `check_convergence()` with the
  thresholds in force when the estimates were extracted. `NA`, not 0,
  when no row carries a verdict, as with a hand-built estimates tibble.

## Error metrics

`bias`
: The mean signed error, $\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)$.

`rmse`
: The root mean squared error,
  $\sqrt{\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)^2}$.

`coverage`
: The share of intervals containing the generating value,
  $\frac{1}{n}\sum_i \mathbb{1}[l_i \le \theta_i \le u_i]$. The interval
  is closed, so a bound equal to the truth counts as covered. Its
  nominal value is the `ci_level` of the estimates, 0.95 by default.

`ci_width`
: The mean interval width, $\frac{1}{n}\sum_i (u_i - l_i)$.

These four are `NA` only when no complete pair exists.

## Correlation metrics

`r`, `r_low`, `r_high`
: The Pearson correlation between estimates and truth with a 95%
  confidence interval from Fisher's $z$ transformation,
  $z = \operatorname{atanh}(r)$ with standard error $1/\sqrt{n-3}$,
  transformed back with $\tanh$. The interval is always 95% and does not
  follow `ci_level`, which is the mass of the posterior intervals.

`rank_r`
: The Spearman rank correlation.

`ccc`
: Lin's concordance correlation coefficient [@linConcordanceCorrelationCoefficient1989],
  computed with population moments (dividing by $n$):
  $$\rho_c = \frac{2 s_{\hat\theta\theta}}{s^2_{\hat\theta} + s^2_\theta + (\bar{\hat\theta} - \bar\theta)^2}
  = 1 - \frac{\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)^2}{s^2_{\hat\theta} + s^2_\theta + (\bar{\hat\theta} - \bar\theta)^2}.$$
  The second form shows it as a rescaled mean squared deviation from
  the identity line.

`ccc_low`, `ccc_high`
: A 95% confidence interval from Lin's $Z$ transformation,
  $Z = \operatorname{atanh}(\rho_c)$, with Lin's asymptotic variance
  for random bivariate normal pairs, which depends on $n$, $r$,
  $\rho_c$ and the location shift
  [@linConcordanceCorrelationCoefficient1989, p. 259, Eq. 2, as corrected
  in @CorrectionNoteConcordance2000, p. 325]. `NA` with three or fewer pairs, at
  $\rho_c = \pm 1$, and when $r = 0$. When the generating values are
  fixed by a design grid rather than drawn, the interval is an
  approximation.

`ccc_accuracy`
: Lin's bias-correction factor
  $C_b = 2 / (v + 1/v + u^2)$, so that $\rho_c = r \cdot C_b$:
  Pearson's $r$ measures precision and $C_b$ accuracy, the distance of
  the best-fitting line from the identity line.

`ccc_scale_shift`
: The ratio of standard deviations, $v = s_{\hat\theta} / s_\theta$.

`ccc_location_shift`
: The mean difference relative to the geometric mean of the two
  standard deviations,
  $u = (\bar{\hat\theta} - \bar\theta) / \sqrt{s_{\hat\theta}\, s_\theta}$.

`calibration_slope`
: The slope of the generating value regressed on the estimate,
  $s_{\hat\theta\theta} / s^2_{\hat\theta} = r / v$. See
  [Reading the concordance](#reading-the-concordance).

`truth_sd`
: The standard deviation of the generating values, $s_\theta$
  (population moments).

All correlation metrics are `NA`, never 0, when fewer than three
complete pairs exist or when estimates or truth have no spread. This is
why population-level correlations are `NA` in a grid whose replications
share their population values. With exactly three pairs, `r` has a point
estimate and no interval.

### Subject level: within replication, then combined

At subject level, each replication is correlated on its own. The
Pearson and Spearman correlations are then combined on Fisher's $z$
scale with weights $n_k - 3$, where $n_k$ is the number of subjects in
replication $k$:

$$\bar z = \frac{\sum_k (n_k - 3)\, z_k}{\sum_k (n_k - 3)}, \qquad \operatorname{SE}(\bar z) = \frac{1}{\sqrt{\sum_k (n_k - 3)}}.$$

`r` is $\tanh(\bar z)$ and its interval is $\tanh(\bar z \pm 1.96\,
\operatorname{SE})$. With one replication this is the ordinary Fisher
interval. `rank_r` is combined the same way, without an interval.

`ccc` is combined on Lin's $Z$ scale, which approaches normality much
faster than $\rho_c$ itself [@linConcordanceCorrelationCoefficient1989,
pp. 259, 268]. The weights are the inverse of each replication's
asymptotic variance $\sigma^2_{Z,k}$ rather than $n_k - 3$, because that
variance also depends on $r$ and the location shift:

$$\bar Z = \frac{\sum_k Z_k / \sigma^2_{Z,k}}{\sum_k 1 / \sigma^2_{Z,k}}, \qquad \operatorname{SE}(\bar Z) = \frac{1}{\sqrt{\sum_k 1/\sigma^2_{Z,k}}}.$$

If any replication has a coefficient but no variance (three subjects,
or a coefficient of exactly 0 or 1), `ccc` is the unweighted mean on
the $Z$ scale and the interval is `NA`. Dropping those replications
would drop exactly the extreme values.

The pooled interval is slightly too narrow when many small replications
are combined. Each replication's $Z$ is a little biased towards zero,
pooling keeps that bias while the standard error shrinks, and in a
simulation of calibrated posterior means with 30 subjects per
replication its coverage was .936 with 5 replications and .928 with 20
(500 studies each).

`ccc_scale_shift` and `calibration_slope` are ratios and are combined
as geometric means, so that $v$ and $1/v$ count as equal departures
from 1. `ccc_accuracy` and `ccc_location_shift` are plain means, and
`truth_sd` is the square root of the mean variance. At subject level
$\rho_c = r \cdot C_b$ therefore holds only approximately.

## Reading the concordance

$r$ asks whether the estimates order the generating values correctly.
$\rho_c$ asks whether they reproduce the values themselves. Which one
matters depends on the question: for individual differences the rank
order is what carries over to correlations with other measures, while
absolute values matter when a parameter is compared across studies or
against a fixed criterion
[@katahiraDoesReliabilityComputational2024, p. 2477].

Hierarchical models complicate the second reading. A posterior mean is
shrunk towards the population mean, and when the model is calibrated
the shrinkage is exactly as large as the data warrant. The error
$\theta - \hat\theta$ is then uncorrelated with $\hat\theta$, so
$\operatorname{cov}(\hat\theta, \theta) = \operatorname{var}(\hat\theta)$, and
three things follow:

- $v = r$: the scale shift equals the recovery correlation, not 1;
- `calibration_slope` $= r / v = 1$;
- with no location shift, $\rho_c = 2r^2 / (1 + r^2)$, a function of
  $r$ alone.

A scale shift below 1 is therefore not a defect by itself.
`calibration_slope` is the column to read against 1: above 1 the
estimates are shrunk more than the data warrant, below 1 less.

<!-- Numbers from a conjugate normal simulation run 2026-09-14, kept with
     the maintainers' development records; slope column is b(truth|est). -->

The table shows this for a normal model with exact posterior means,
50 subjects and 2000 replications per row. The unpooled estimate is the
per-subject maximum likelihood estimate.

| generating model and prior | estimator | $r$ | $\rho_c$ | $v$ | slope | $u$ | RMSE |
|---|---|---|---|---|---|---|---|
| calibrated, reliability .50 | unpooled | .704 | .657 | 1.420 | 0.50 | −.002 | .997 |
| calibrated, reliability .50 | posterior mean | .704 | .657 | 0.710 | 1.00 | −.005 | .706 |
| prior SD too small | posterior mean | .705 | .365 | 0.284 | 2.50 | .014 | .820 |
| prior SD too large | posterior mean | .704 | .691 | 1.137 | 0.63 | .000 | .819 |
| prior mean off by 0.5 | posterior mean | .703 | .629 | 0.712 | 1.00 | .306 | .748 |

Because $C_b$ treats $v$ and $1/v$ alike, $\rho_c$ cannot tell
calibrated shrinkage from no shrinkage (.657 in both rows), and it
*rewards* too little shrinkage (.691 against .657) although the RMSE
is worse. It does detect too much shrinkage and a shifted location.
Read `ccc` together with `rmse` and `calibration_slope`, and use `r`
when only the rank order of subjects matters.

The table above is a conjugate toy, where the posterior means are exact
and the algebra therefore holds to the last decimal.
[Hierarchical estimation against subject-wise maximum
likelihood](hierarchical-vs-ml.html) is the same contrast for a real bmm
model, sampled rather than derived: `mixture2p` fitted hierarchically
and subject by subject on one data set, and again over a grid.

Two caveats. The identities above assume that the generating values
are drawn from the population the model estimates; with generating
values from a hand-picked uniform range they hold only roughly. And
$\rho_c$, like $r$, grows with the spread of the generating values at a
fixed measurement error: in the same simulation, doubling `truth_sd`
raised $\rho_c$ for the calibrated posterior mean from .657 to .885.
Compare either coefficient only between rows with similar `truth_sd`.

## The scale

`recover()` and `recover_subjects()` score on the natural scale by
default: estimates, interval bounds and truth are converted with
`inverse_link()` before any metric is computed, and `scale = "link"`
skips the conversion. Coverage does not depend on the choice, because
a monotone transformation keeps a value inside its interval, but bias,
RMSE, interval width and the correlations do.

Two links need care at zero. Under the `sqrt` link, an interval that
spans zero on the link scale becomes $[0, \max(l^2, u^2)]$. Under the
`inverse` link, an interval that spans zero has no image on the natural
scale, so its bounds and `covered` are `NA` and a warning names the
term.

## Column names

The first nine columns of a recovery object (`term`, `estimate`,
`ci_low`, `ci_high`, `ci_method`, `ci_level`, `rhat`, `ess_bulk`,
`ess_tail`) follow the column contract of the
[apabayes](https://github.com/GidonFrischkorn/apabayes) package, so
recovery tables can be formatted there without renaming.

## References
