---
title: "Recovery summary columns"
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
library(bmmtools)
```

This page defines every column `summary()` returns for a
`bmmtools_recovery` object: what is computed, over which rows, and when
the value is `NA`. It describes the computation only.

```{r}
s <- tibble::as_tibble(summary(recovery_mixture2p))
print(s[1:4, c(
  "condition", "term", "level", "scale", "n", "n_replications",
  "n_converged", "bias", "rmse", "coverage", "coverage_50", "ci_width",
  "ci_width_50"
)], width = Inf)
print(s[1:4, c(
  "condition", "term", "level", "r", "r_low", "r_high", "rank_r", "ccc",
  "calibration_slope", "calibration_intercept"
)], width = Inf)
print(s[1:4, c(
  "condition", "term", "level", "z_mean", "z_sd", "z_mean_mcse",
  "z_sd_mcse", "contraction"
)], width = Inf)
```

## Which rows each row summarises

One summary row covers one parameter (`term`) at one `level`, and one
`condition` when the object comes from `recovery_grid()`. Write
$\hat\theta_i$ for the estimate in row $i$ (the posterior median),
$\theta_i$ for the generating value, and $[l_i, u_i]$ for the credible
interval, all on the scale in the `scale` column.

At the **population** level the rows are the replications, one estimate
per fit. At the **subject** level the rows are subjects within
replications; the error metrics use all of them, and the correlation
metrics are computed within each replication and then combined, as
described below.

Missing values are dropped pairwise before anything is counted, so `n`
always reports the pairs a metric used.

## Identification columns

`term`, `level`, `scale`
: The parameter, `"population"` or `"subject"`, and `"natural"` or
  `"link"`.

`n`
: Complete estimate–truth pairs. At subject level, the mean number of
  subjects per replication.

`n_replications`
: Distinct replications contributing to the row.

`n_converged`
: Replications whose fit passed `check_convergence()` with the
  thresholds in force when the estimates were extracted. `NA`, not 0,
  when no row carries a verdict, as with a hand-built estimates tibble.

## Error metrics

`bias`
: The mean signed error, $\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)$.

`rmse`
: The root mean squared error,
  $\sqrt{\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)^2}$.

`coverage`
: The share of intervals containing the generating value,
  $\frac{1}{n}\sum_i \mathbb{1}[l_i \le \theta_i \le u_i]$. The interval
  is closed, so a bound equal to the truth counts as covered. Its
  nominal value is the `ci_level` of the estimates, 0.95 by default.

`ci_width`
: The mean interval width, $\frac{1}{n}\sum_i (u_i - l_i)$.

`coverage_50`, `ci_width_50`
: The same two for the central 50 % interval, the 25th to the 75th
  percentile of the draws, stored beside the `ci_level` interval as
  `ci_low_50` and `ci_high_50`. Its nominal coverage is 0.5 whatever
  `ci_level` is. Where the 95 % interval checks the tails of the
  posterior, this one checks its centre: a coverage of 0.95 with a
  `coverage_50` far above 0.5 says the intervals are too wide in the
  middle. `NA` when the estimates carry no inner bounds, as in a
  hand-built estimates tibble.

These are `NA` only when no complete pair exists, except for the inner
columns as just described.

## Correlation metrics

`r`, `r_low`, `r_high`
: The Pearson correlation between estimates and truth with a 95%
  confidence interval from Fisher's $z$ transformation,
  $z = \operatorname{atanh}(r)$ with standard error $1/\sqrt{n-3}$,
  transformed back with $\tanh$. The interval is always 95% and does not
  follow `ci_level`, which is the mass of the posterior intervals.

`rank_r`
: The Spearman rank correlation.

`ccc`
: Lin's concordance correlation coefficient [@linConcordanceCorrelationCoefficient1989],
  computed with population moments (dividing by $n$):
  $$\rho_c = \frac{2 s_{\hat\theta\theta}}{s^2_{\hat\theta} + s^2_\theta + (\bar{\hat\theta} - \bar\theta)^2}
  = 1 - \frac{\frac{1}{n}\sum_i (\hat\theta_i - \theta_i)^2}{s^2_{\hat\theta} + s^2_\theta + (\bar{\hat\theta} - \bar\theta)^2}.$$
  The second form shows it as a rescaled mean squared deviation from
  the identity line.

`ccc_low`, `ccc_high`
: A 95% confidence interval from Lin's $Z$ transformation,
  $Z = \operatorname{atanh}(\rho_c)$, with Lin's asymptotic variance
  for random bivariate normal pairs, which depends on $n$, $r$,
  $\rho_c$ and the location shift
  [@linConcordanceCorrelationCoefficient1989, p. 259, Eq. 2, as corrected
  in @CorrectionNoteConcordance2000, p. 325]. `NA` with three or fewer pairs, at
  $\rho_c = \pm 1$, and when $r = 0$. When the generating values are
  fixed by a design grid rather than drawn, the interval is an
  approximation.

`ccc_accuracy`
: Lin's bias-correction factor
  $C_b = 2 / (v + 1/v + u^2)$, so that $\rho_c = r \cdot C_b$:
  Pearson's $r$ measures precision and $C_b$ accuracy, the distance of
  the best-fitting line from the identity line.

`ccc_scale_shift`
: The ratio of standard deviations, $v = s_{\hat\theta} / s_\theta$.

`ccc_location_shift`
: The mean difference relative to the geometric mean of the two
  standard deviations,
  $u = (\bar{\hat\theta} - \bar\theta) / \sqrt{s_{\hat\theta}\, s_\theta}$.

`calibration_slope`
: The slope of the generating value regressed on the estimate,
  $s_{\hat\theta\theta} / s^2_{\hat\theta} = r / v$. See
  [Reading the concordance](#reading-the-concordance).

`calibration_intercept`
: The intercept of the same regression,
  $a = \bar\theta - b\,\bar{\hat\theta}$ with $b$ the calibration
  slope: the generating value the line predicts at an estimate of 0. It
  is read on the scale of the row, so a summary on the link scale and
  one on the natural scale give different intercepts for the same fits.
  It is not calibration-in-the-large, $\bar\theta - \bar{\hat\theta}$,
  which fixes the slope at 1 and equals $-$`bias`. The two are related
  by $a = (\bar\theta - \bar{\hat\theta}) - (b - 1)\,\bar{\hat\theta}$,
  so when the slope is far from 1 and 0 lies far from the estimates, the
  intercept mostly restates the slope. Read the two together, as one
  line.

`truth_sd`
: The standard deviation of the generating values, $s_\theta$
  (population moments).

All correlation metrics are `NA`, never 0, when fewer than three
complete pairs exist or when estimates or truth have no spread. This is
why population-level correlations are `NA` in a grid whose replications
share their population values. With exactly three pairs, `r` has a point
estimate and no interval.

### Subject level: within replication, then combined

At subject level, each replication is correlated on its own. The
Pearson and Spearman correlations are then combined on Fisher's $z$
scale with weights $n_k - 3$, where $n_k$ is the number of subjects in
replication $k$:

$$\bar z = \frac{\sum_k (n_k - 3)\, z_k}{\sum_k (n_k - 3)}, \qquad \operatorname{SE}(\bar z) = \frac{1}{\sqrt{\sum_k (n_k - 3)}}.$$

`r` is $\tanh(\bar z)$ and its interval is $\tanh(\bar z \pm 1.96\,
\operatorname{SE})$. With one replication this is the ordinary Fisher
interval. `rank_r` is combined the same way, without an interval.

`ccc` is combined on Lin's $Z$ scale, which approaches normality much
faster than $\rho_c$ itself [@linConcordanceCorrelationCoefficient1989,
pp. 259, 268]. The weights are the inverse of each replication's
asymptotic variance $\sigma^2_{Z,k}$ rather than $n_k - 3$, because that
variance also depends on $r$ and the location shift:

$$\bar Z = \frac{\sum_k Z_k / \sigma^2_{Z,k}}{\sum_k 1 / \sigma^2_{Z,k}}, \qquad \operatorname{SE}(\bar Z) = \frac{1}{\sqrt{\sum_k 1/\sigma^2_{Z,k}}}.$$

If any replication has a coefficient but no variance (three subjects,
or a coefficient of exactly 0 or 1), `ccc` is the unweighted mean on
the $Z$ scale and the interval is `NA`. Dropping those replications
would drop exactly the extreme values.

The pooled interval is slightly too narrow when many small replications
are combined. Each replication's $Z$ is a little biased towards zero,
pooling keeps that bias while the standard error shrinks, and in a
simulation of calibrated posterior means with 30 subjects per
replication its coverage was .936 with 5 replications and .928 with 20
(500 studies each).

`ccc_scale_shift` and `calibration_slope` are ratios and are combined
as geometric means, so that $v$ and $1/v$ count as equal departures
from 1. `ccc_accuracy`, `ccc_location_shift` and
`calibration_intercept` are plain means (an intercept can be 0 or
negative), and `truth_sd` is the square root of the mean variance. At subject level
$\rho_c = r \cdot C_b$ therefore holds only approximately.

## z-score and contraction

These two are computed on the link scale whatever the `scale` column
says, from the posterior mean $m_i$ and posterior standard deviation
$s_i$ that each row stores as `post_mean_link` and `post_sd_link`. For a
maximum likelihood row, $m_i$ is the point estimate and $s_i$ its
standard error. [Coverage is not recovery](#coverage-is-not-recovery)
shows how to read them.

`z_mean`, `z_sd`
: The mean and standard deviation of the posterior z-scores
  $z_i = (m_i - \theta_i) / s_i$, one per row. Because each $z_i$ is
  already standardised, the subject rows of all replications are pooled
  directly, without the within-then-combine step of the correlations.
  Rows whose $z_i$ is `NA` are left out.

`z_mean_mcse`, `z_sd_mcse`
: Their Monte Carlo standard errors, $\operatorname{SD}(z)/\sqrt{n}$ and
  $\operatorname{SD}(z)/\sqrt{2(n-1)}$. For independent, normally
  distributed $z_i$ the first is exact and the second the large-sample
  approximation to the sampling error of an SD. Subject rows of one
  replication share a population posterior, and heavy tails widen the sampling error of a
  standard deviation, so at subject level both understate it.

`contraction`
: The mean, over the rows that have one, of
  $c_i = 1 - s_i^2 / \sigma^2_{\text{prior},i}$,
  the share of the prior variance that the data removed. It needs the
  prior standard deviation of each parameter, given as `prior_sd` to
  `recover()` or `recovery_grid()`: a table such as
  `prior_sd_table()` returns, or a prior-only fit. Without one it is
  `NA`, as in `recovery_mixture2p` above. It is also `NA` at subject
  level, where a posterior narrows through pooling as well as through
  the subject's own data, and for maximum likelihood rows, which have
  no prior. It is not clamped at 0: a posterior wider than its prior is
  a finding, not a rounding error.

## Reading the concordance

$r$ asks whether the estimates order the generating values correctly.
$\rho_c$ asks whether they reproduce the values themselves. Which one
matters depends on the question: for individual differences the rank
order is what carries over to correlations with other measures, while
absolute values matter when a parameter is compared across studies or
against a fixed criterion
[@katahiraDoesReliabilityComputational2024, p. 2477].

Hierarchical models complicate the second reading. A posterior mean is
shrunk towards the population mean, and when the model is calibrated
the shrinkage is exactly as large as the data warrant. The error
$\theta - \hat\theta$ is then uncorrelated with $\hat\theta$, so
$\operatorname{cov}(\hat\theta, \theta) = \operatorname{var}(\hat\theta)$, and
three things follow:

- $v = r$: the scale shift equals the recovery correlation, not 1;
- `calibration_slope` $= r / v = 1$;
- with no location shift, $\rho_c = 2r^2 / (1 + r^2)$, a function of
  $r$ alone.

A scale shift below 1 is therefore not a defect by itself.
`calibration_slope` is the column to read against 1: above 1 the
estimates are shrunk more than the data warrant, below 1 less.

<!-- Numbers from a conjugate normal simulation run 2026-09-14, kept with
     the maintainers' development records; slope column is b(truth|est). -->

The table shows this for a normal model with exact posterior means,
50 subjects and 2000 replications per row. The unpooled estimate is the
per-subject maximum likelihood estimate.

| generating model and prior | estimator | $r$ | $\rho_c$ | $v$ | slope | $u$ | RMSE |
|---|---|---|---|---|---|---|---|
| calibrated, reliability .50 | unpooled | .704 | .657 | 1.420 | 0.50 | −.002 | .997 |
| calibrated, reliability .50 | posterior mean | .704 | .657 | 0.710 | 1.00 | −.005 | .706 |
| prior SD too small | posterior mean | .705 | .365 | 0.284 | 2.50 | .014 | .820 |
| prior SD too large | posterior mean | .704 | .691 | 1.137 | 0.63 | .000 | .819 |
| prior mean off by 0.5 | posterior mean | .703 | .629 | 0.712 | 1.00 | .306 | .748 |

Because $C_b$ treats $v$ and $1/v$ alike, $\rho_c$ cannot tell
calibrated shrinkage from no shrinkage (.657 in both rows), and it
*rewards* too little shrinkage (.691 against .657) although the RMSE
is worse. It does detect too much shrinkage and a shifted location.
Read `ccc` together with `rmse` and `calibration_slope`, and use `r`
when only the rank order of subjects matters.

The table above is a conjugate toy, where the posterior means are exact
and the algebra therefore holds to the last decimal.
[Hierarchical estimation against subject-wise maximum
likelihood](hierarchical-vs-ml.html) is the same contrast for a real bmm
model, sampled rather than derived: `mixture2p` fitted hierarchically
and subject by subject on one data set, and again over a grid.

Two caveats. The identities above assume that the generating values
are drawn from the population the model estimates; with generating
values from a hand-picked uniform range they hold only roughly. And
$\rho_c$, like $r$, grows with the spread of the generating values at a
fixed measurement error: in the same simulation, doubling `truth_sd`
raised $\rho_c$ for the calibrated posterior mean from .657 to .885.
Compare either coefficient only between rows with similar `truth_sd`.

## Coverage is not recovery

An interval can contain the generating value for two reasons: the data
located the parameter, or the prior put its mass near the truth.
Coverage counts both alike. In a recovery study the analyst chooses the
generating values and the priors, usually from the same sense of what
is plausible, so the second reason is easy to meet by accident. The
case that motivated these columns came from the development of a
linear ballistic accumulator model for bmm, not from bmmtools: a
recovery reported every parameter inside its 95 % interval, while in
one scenario the likelihood was flat along one direction of the
parameter space. Along that direction the posterior SD was 0.494 and
the prior SD 0.500, a contraction of $1 - 0.494^2 / 0.500^2 = 0.024$.
The interval covered because the prior did.

Contraction and the z-score separate the two reasons
[@schadPrincipledBayesianWorkflow2020]. Contraction says how much the
data narrowed the prior: near 1 the posterior is the data's, near 0 it
is still the prior. The z-score says how far the posterior mean is
from the truth in units of posterior SD. Plotted against each other,
they mark four regions [@schadPrincipledBayesianWorkflow2020, Figure 5]:
high contraction with small $|z|$ is the ideal; high contraction with
large $|z|$ is a posterior fitted to noise in the data; low contraction
with small $|z|$ is a parameter the design does not identify; and low
contraction with large $|z|$ is a prior in conflict with the likelihood.

```{r contraction-data, include = FALSE}
ct <- readRDS(from_root("vignettes/articles/assets/metrics-contraction.rds"))
rows <- tibble::as_tibble(ct$out)
rows$n_trials <- ct$cells$n_trials[match(rows$condition, ct$cells$condition)]
ct_summary <- tibble::as_tibble(ct$summary)
ct_summary$n_trials <- ct$cells$n_trials[
  match(ct_summary$condition, ct$cells$condition)
]
pick <- function(x, trials, term, level) {
  x[x$n_trials == trials & x$term == term & x$level == level, ]
}
one_sd <- pick(rows, 1, "thetat", "sd")
sum_one_sd <- pick(ct_summary, 1, "thetat", "sd")
sum_hundred_sd <- pick(ct_summary, 100, "thetat", "sd")
one_pop <- rows[rows$n_trials == 1 & rows$level == "population", ]
one_sd_rows <- rows[rows$n_trials %in% c(1, 3) & rows$level == "sd", ]
hundred <- rows[rows$n_trials == 100, ]
# the prose below states these; a regenerated asset must still satisfy them
stopifnot(all(one_sd$covered), all(one_sd_rows$z > 0))
```

A design that informs every parameter well puts all its points in the
ideal corner, which shows little. The grid below is built to reach the
low end as well: `mixture2p` with 20 subjects and 1, 3, 10, 30 or 100
trials each, `r max(rows$replication)` replications per row, bmm's default
priors, and their closed-form SDs through `prior_sd = "analytic"`.

```{r contraction-run, eval = FALSE, code = shown("run", "vignettes/articles/precompute/metrics-contraction.R")}
```

```{r contraction-plot, echo = FALSE, fig.height = 3.8}
library(ggplot2)
ggplot(rows, aes(contraction, z, colour = factor(n_trials))) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point(size = 1.8, alpha = 0.8) +
  facet_grid(level ~ term) +
  scale_colour_viridis_d(end = 0.9) +
  labs(
    x = "Contraction", y = "Posterior z-score",
    colour = "Trials per\nsubject"
  )
```

`r sum(rows$covered)` of the `r nrow(rows)` intervals contain the
generating value, so a coverage table finds nothing wrong with this
grid. (The count includes the one replication, at 10 trials, whose fit
missed bmmtools' convergence thresholds.) Contraction does. With 100 trials per subject it lies
between `r fmt(min(hundred$contraction), 3)` and
`r fmt(max(hundred$contraction), 3)` for every term. With one trial it
falls to `r fmt(min(one_pop$contraction))`--`r fmt(max(one_pop$contraction))`
for the population values, and for the SD of `thetat` to a median of
`r fmt(stats::median(one_sd$contraction))`; in
`r sum(one_sd$contraction < 0)` replications the posterior is wider
than the prior (contraction down to `r fmt(min(one_sd$contraction), 3)`).
All `r nrow(one_sd)` of those intervals covered, and their z-scores lie
between `r fmt(min(one_sd$z))` and `r fmt(max(one_sd$z))`. The
replications agree with each other because each reports the prior, not
because each locates the parameter.

### Reading z when the generating values are fixed

The usual reading of z, a mean of 0 for no bias and an SD of 1 for
calibrated uncertainty, holds when the generating values are drawn from
the prior, as in simulation-based calibration and in the procedure of
@schadPrincipledBayesianWorkflow2020. A recovery grid holds them fixed,
and then z also carries the prior's pull. In a conjugate normal model
with prior $\mathcal{N}(\mu_0, \tau^2)$, a fixed truth $\theta$ and
contraction $c$, the z-score over replications is distributed as

$$z \sim \mathcal{N}\!\left(\sqrt{1 - c}\,\delta,\; c\right), \qquad \delta = \frac{\mu_0 - \theta}{\tau},$$

so its SD is $\sqrt{c}$, the posterior SD over the standard error of
the data-only estimate, which is below 1. Its mean moves away from 0 on the side of the prior
mean as $c$ falls, because the posterior mean is pulled towards the
prior mean. With
$\theta$ drawn from the prior, the same model gives $z \sim
\mathcal{N}(0, 1)$ at every contraction.

bmm's models are not conjugate, and the formula does not predict the
grid's numbers term by term. The direction shows all the same. The
spread of z shrinks as contraction falls: for the SD of `thetat`, `z_sd` is
`r fmt(sum_hundred_sd$z_sd)` at 100 trials and
`r fmt(sum_one_sd$z_sd, 3)` at one. And every z of the two SD terms at
one and three trials is positive (the smallest is
`r fmt(min(one_sd_rows$z), 3)`): bmm's default prior on an SD,
`exponential(1)`, has its mean at 1, above the generating SDs of
`r ct$sds[["kappa"]]` and `r ct$sds[["thetat"]]`.

Read `z_mean` and `z_sd` against 0 and 1 only where contraction is near
1. At $c = 0.99$ the conjugate shift is $0.1\,\delta$ and the SD
$0.995$. Below that, read `contraction` first. With
`r max(rows$replication)` replications per row, `z_sd_mcse` is
$1/\sqrt{18} \approx 0.24$ of `z_sd` itself, so this grid supports the
direction of these differences, not their size.

## The scale

`recover()` and `recover_subjects()` score on the natural scale by
default: estimates, interval bounds and truth are converted with
`inverse_link()` before any metric is computed, and `scale = "link"`
skips the conversion. Coverage does not depend on the choice, because
a monotone transformation keeps a value inside its interval, but bias,
RMSE, interval width and the correlations do.

Two links need care at zero. Under the `sqrt` link, an interval that
spans zero on the link scale becomes $[0, \max(l^2, u^2)]$. Under the
`inverse` link, an interval that spans zero has no image on the natural
scale, so its bounds and `covered` are `NA` and a warning names the
term.

## Column names

The first nine columns of a recovery object (`term`, `estimate`,
`ci_low`, `ci_high`, `ci_method`, `ci_level`, `rhat`, `ess_bulk`,
`ess_tail`) follow the column contract of the
[apabayes](https://github.com/GidonFrischkorn/apabayes) package, so
recovery tables can be formatted there without renaming.

## References
