---
title: "Validating a model without a built-in adapter"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
library(bmmtools)
res <- readRDS(from_root("vignettes/articles/assets/own-model.rds"))
```

bmmtools ships generator adapters for six bmm models: `sdt_yn`,
`sdt_mafc`, the three-parameter `ezdm`, `ddm`, `mixture2p` and `sdm`.
A model you are developing, for example one started with
`bmm::use_model_template()`, has none, and `simulate_recovery()` stops
with an error asking for `generator`. This article shows how to write
one, what the rest of the loop needs from it, and how to score fits and
estimates that did not come from the generate layer at all.

The example is bmm's three-parameter mixture model, `mixture3p`, which
has no adapter. The code was run once and its results saved, so this
page was built without Stan.

## What a generator is

A generator is a function `function(pars, n_trials, model)` that
returns the rows of **one subject** as a data frame, with the column
names the model expects. `simulate_recovery()` calls it once per
subject and adds the `id` column itself.

- `pars` is a named list of that subject's parameter values on the
  **natural scale**, fixed parameters included.
- `n_trials` is the number of trials per subject, as given to
  `simulate_recovery()` or in a grid row.
- `model` is the model object, so the generator can read column names
  or other settings from it.

`simulate_recovery()` takes `pars` on the link scale, draws subject
values there and converts them with `inverse_link()` using the model's
link table before calling the generator. Check that table first:

```{r model, eval = FALSE, code = shown("model", "vignettes/articles/precompute/own-model.R")}
```

```{r, echo = FALSE}
res$links
```

`kappa` arrives in the generator as a precision, but `thetat` and
`thetant` have an identity link: in `mixture3p` they are softmax weights
against a guessing weight fixed at 0. The generator has to turn them
into probabilities itself.

## Writing the generator

```{r generator, eval = FALSE, code = shown("generator", "vignettes/articles/precompute/own-model.R")}
```

The non-target locations are drawn inside the generator, so they are
part of the simulated data. `simulate_recovery()` applies its `seed`
around the generator too, so these draws are reproducible with the
rest. The generator's output is checked against the model's response
columns before anything is fitted.

## Simulate, fit and score

From here the loop is the same as for a model with an adapter:

```{r simulate, eval = FALSE, code = shown("simulate", "vignettes/articles/precompute/own-model.R")}
```

```{r, echo = FALSE}
res$data
res$truth$subjects
```

`thetant` is in `pars` but not in `sds`, so it does not vary between
subjects and has no rows in `truth$subjects`: a parameter that does not
vary has nothing person-level to recover.

```{r fit, eval = FALSE, code = shown("fit", "vignettes/articles/precompute/own-model.R")}
```

```{r, echo = FALSE}
res$convergence
```

```{r score, eval = FALSE, code = shown("score", "vignettes/articles/precompute/own-model.R")}
```

```{r, echo = FALSE}
res$population
summary(res$subjects)
```

The same generator runs a whole design through `recovery_grid()`:

```{r, eval = FALSE}
recovery_grid(
  model,
  grid = expand.grid(n_subjects = c(30, 60), n_trials = c(50, 100)),
  pars = c(kappa = log(8), thetat = 1.5, thetant = 0),
  sds = c(kappa = 0.3, thetat = 0.5),
  generator = generate_mixture3p,
  dir = "fits/mixture3p", reps = 10, seed = 1
)
```

## Scoring estimates you already have

`recover()` and `recover_subjects()` accept an **estimates tibble** in
place of a fit. That is how to score estimates from another fitting
routine, from an older simulation, or from any pipeline where no fit
object is at hand. No fitting package needs to be installed.

The tibble needs the columns `term`, `estimate`, `ci_low`, `ci_high`,
`ci_method`, `ci_level`, `rhat`, `ess_bulk`, `ess_tail`, `level` and
`id`, and optionally `converged` and `replication`. Estimates and truth
are on the link scale; `links` maps each term to its link so that
scoring can happen on the natural scale:

```{r}
estimates <- tibble::tibble(
  term = rep(c("kappa", "thetat"), times = 3),
  estimate = c(2.10, 1.00, 1.70, 0.60, 2.45, 1.35),
  ci_low = c(1.60, 0.40, 1.20, 0.05, 1.95, 0.80),
  ci_high = c(2.60, 1.60, 2.20, 1.15, 2.95, 1.90),
  ci_method = "eti", ci_level = 0.95,
  rhat = 1, ess_bulk = 900, ess_tail = 900,
  level = "population", id = NA_character_,
  replication = rep(1:3, each = 2)
)
truth <- tibble::tibble(
  term = rep(c("kappa", "thetat"), times = 3),
  true_value = c(2.0, 1.1, 1.8, 0.5, 2.4, 1.4),
  replication = rep(1:3, each = 2)
)

recovered <- recover(estimates, truth, links = c(kappa = "log", thetat = "logit"))
summary(recovered)
```

`n_converged` is `NA` because the tibble carries no convergence verdict.
When `truth` has a `replication` column it is used as a join key; without
one, the same truth is scored against every replication.

`extract_estimates()` produces this tibble from a fit, so a scoring
pipeline can be written and tested on saved tibbles and pointed at fits
later.

## Scoring a brms fit

Scoring dispatches on `brmsfit`, and a `bmmfit` is one, so a plain
brms fit can be scored as well. Without bmm's link table, scoring falls
back to the link scale and says so, unless `links` is given:

```{r, eval = FALSE}
recover(brms_fit, truth, links = c(kappa = "log", thetat = "logit"))
```

Term names are the coefficient names with the `b_` prefix and the
coefficient suffix removed, so `b_kappa_Intercept` becomes `kappa`.
`extract_estimates(brms_fit)$term` shows what `truth` has to match.
A term in `truth` that the fit did not estimate is dropped with a
warning that lists the available terms.

## Limits of this version

- Each parameter is scored through its intercept, or through one cell
  mean per task (`kappa ~ 0 + task`, the terms `kappa_task1`,
  `kappa_task2`, ...), which is what `recovery_formula()` writes with and
  without `task_col`. A formula with an intercept and other coefficients
  for one parameter, say `kappa ~ 1 + setsize`, stops with an error
  rather than join the estimates wrongly.
- Group-level standard deviations are scored on the link scale only
  (`recover(level = "sd")`). Group-level correlations are extracted
  (`extract_estimates(level = "cor")`) but not yet scored.
- `prior_check()` derives the floor and ceiling of the response only
  for the models with an adapter; for your model, pass `range` or a
  `summary` function of your own.

## Contributing an adapter

If your model is part of bmm and its generator fits the pattern above,
an adapter in bmmtools saves every user from writing it. Adapters live
in `R/adapters.R`: a function per model that maps one subject's
natural-scale parameters onto the model's `r*()` function and names the
output columns from the model object, plus an entry in the table that
looks them up by class. A response range for `prior_check()` goes in
`response_range()` in `R/prior-check.R`. Open an issue or a pull request
on [GitHub](https://github.com/GidonFrischkorn/bmmtools).
