---
title: "Prior predictive checks on the observable scale"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
library(bmmtools)
res <- readRDS(from_root("vignettes/articles/assets/prior-check.rds"))
pct <- function(set, statistic) {
  x <- prior_check_sdt_yn
  value <- x$value[x$prior == set & x$statistic == statistic]
  sprintf("%.1f%%", 100 * value)
}
```

Priors of cognitive measurement models are set on the link scale, and
what a `normal(0, 1)` there means for the data is rarely obvious.
`prior_check()` answers that directly: it fits the model with
`sample_prior = "only"`, draws from the prior predictive distribution
and summarises the draws on the scale of the response. Because the
draws come from bmm's own fit, the prior being checked is exactly the
prior bmm fits with, defaults included.

The fits go through `fit_cached()`, so each prior set compiles and
samples once and later calls reuse it. Two checks are shown below: a
signal detection model, whose count response has a floor and a ceiling,
and a mixture model with a statistic of our own.

## A count response: sdt_yn

The first example is the check behind `prior_check_sdt_yn`, the prior
check that ships with the package. It compares bmm's default priors with
a set that narrows the priors on the between-subject SDs:

```{r prior-check, eval = FALSE, code = shown("prior-check", "data-raw/example-objects.R")}
```

`prior` is either one prior (or `NULL` for bmm's defaults) or a **named
list** of prior sets. Each set is fitted separately, its fit cached at
`<file>-<set>`, and becomes one level of the `prior` column.

```{r}
prior_check_sdt_yn
```

The default summary has five statistics per response:

- `floor_rate` and `ceiling_rate`: the share of all prior-predictive
  values, over draws and observations, at or below the floor and at or
  above the ceiling of the response;
- `q50`, `q90` and `q95`: quantiles of the prior-predictive values.

For `sdt_yn` and `sdt_mafc` the floor is 0 and the ceiling is the trial
count of each row, read from the model's `n_trials` column. With
exactly two prior sets, `summary()` adds a `difference` column, the
second set minus the first. Here, `r pct("default", "floor_rate")` of
the prior-predictive counts under bmm's defaults are at 0 and
`r pct("default", "ceiling_rate")` at the ceiling; with the narrower SD
priors the shares are `r pct("narrow_sd", "floor_rate")` and
`r pct("narrow_sd", "ceiling_rate")`.

```{r}
summary(prior_check_sdt_yn)
```

For a discrete response, plot the draws as histograms:

```{r}
plot_prior_check(prior_check_sdt_yn, type = "histogram")
```

`type = "statistic"` plots the summary table instead, one point per
statistic and prior set:

```{r}
plot_prior_check(prior_check_sdt_yn, type = "statistic")
```

## A continuous response and a statistic of your own

For a continuous response, a value exactly at a bound has probability
zero, so the two rates are only informative for counts. For the
circular models `mixture2p` and `sdm` the range is set to `-pi` and
`pi`, and the rates report mass outside it. The quantiles of a signed
circular error are symmetric about zero, so a statistic on the absolute
error says more. `summary` takes any function of `yrep`, the matrix of
prior-predictive draws (draws in rows, observations in columns), and
`data`, and returns a data frame with the columns `statistic` and
`value`, plus `response` for a model with several responses:

```{r mixture-check, eval = FALSE, code = shown("mixture-check", "vignettes/articles/precompute/prior-check.R")}
```

```{r, echo = FALSE}
summary(res$mix_check)
```

The default plot draws one thin density line per prior-predictive draw,
so the spread between draws stays visible, with the observed response
from `data` on top:

```{r mixture-plot, eval = FALSE, code = shown("mixture-plot", "vignettes/articles/precompute/prior-check.R")}
```

```{r, echo = FALSE}
plot_prior_check(res$mix_check, draws = 30)
```

`observed = FALSE` removes the overlay. It is left out without comment
when `data` has no response column, for example when the check runs
over a design skeleton rather than simulated or real data.

## Arguments worth knowing

`range`
: The floor and the ceiling of the response as two numbers. `NULL`
  derives them for `sdt_yn`, `sdt_mafc`, `mixture2p` and `sdm`; for any
  other model the two rates are `NA` and a message says so.

`n_draws`
: Prior-predictive draws kept per prior set. Asking for more than the
  fit holds is capped with a message rather than refused.

`seed`
: Passed to the fitter, where it enters the cache key, and used around
  the draw subsample, so the same seed keeps the same draws.

`file`
: Where the prior-only fits are cached. `NULL` uses a temporary file, so
  the compiled model is reused within the session and nothing is left
  on disk.

`...`
: Passed to `fit_cached()` and on to `bmm::bmm()`: `chains`, `iter`,
  `backend`, `cores`. Giving `sample_prior` is an error, because the
  check sets it.

A population-level slope without a proper prior triggers a warning
before fitting, because prior-only sampling has nothing to draw a flat
prior from. bmm's defaults cover the intercepts and the group-level
SDs, so the warning concerns slopes you added to the formula.

## What the object carries

Besides the summary rows, a `bmmtools_prior_check` keeps the draws, the
`brms::prior_summary()` of each fit, the data, the model, the formula
and the cache files as attributes. The prior summary records the prior
that produced the draws, which is the one to report:

```{r}
attr(prior_check_sdt_yn, "priors")$narrow_sd
```
