---
title: "Running a recovery study"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
library(bmmtools)
```

A recovery study repeats the loop of
[Validating a new bmm model](bmmtools.html), simulate, fit and score,
over a design: several numbers of subjects and trials, several
replications of each. `recovery_grid()` runs that loop and takes care of
what makes long simulations fragile: every cell is saved as soon as it
finishes, an interrupted run resumes from those files, and a short
preflight fit catches a broken model before hours are spent on it.

The example below is the grid behind `recovery_mixture2p`, the recovery
object that ships with the package. Its results are used live on this
page. The run took 8 minutes 58 seconds, compilation and preflight
included.

## The call

```{r grid, eval = FALSE, code = shown("grid", "data-raw/example-objects.R")}
```

Everything after `seed` is passed to `fit_cached()` and from there to
`bmm::bmm()`.

## The design grid

`grid` is a data frame with one row per condition. It needs the columns
`n_subjects` and `n_trials`; any other column named after a parameter
sets that parameter's population value for the row, and a column
`sd_<parameter>` sets its between-subject SD. Both override `pars` and
`sds`, which act as the defaults for every row:

```{r, eval = FALSE}
grid <- expand.grid(
  n_subjects = c(20, 50),
  n_trials = c(30, 100),
  kappa = log(c(4, 16))
)
```

A design created with `SimDesign::createDesign()` is a data frame and
can be passed as it is. `model` can also be a function of one grid row
that returns a model, for designs where the model's arguments change
between rows; the link table must stay the same across rows, because
one table scores every cell.

## Smoke run and preflight

Before the full grid, run it with `smoke = TRUE`. This fits the first
two rows with two replications into `<dir>/smoke`, so a smoke run never
overwrites or gets mistaken for a full one.

By default, the grid also runs a **preflight**: before any cell, the
first cell is fitted once with one chain and 200 iterations into
`<dir>/preflight`. A compile error, an init failure or data the model
rejects stop the grid there, with nothing else run. The preflight is
skipped when the first cell already has a cached fit.

## Files on disk and resuming

Each cell writes two files as soon as it is done, named after its grid
row and replication:

- `cell-<row>-rep-<rep>-sim.rds` holds the simulated data and truth;
- `cell-<row>-rep-<rep>.rds` holds the fit, with its cache key in
  `cell-<row>-rep-<rep>.key`.

To resume after an interruption, run the same call again. A cell whose
simulation file exists reads it instead of simulating, and a cell whose
fit matches its key reuses the fit, so finished cells cost only the
time to read them. Rerunning the script that builds the example
objects, with all 20 grid fits on disk, took 23 seconds, including two
prior-only fits that were not cached yet. If an argument that enters the key
changes, say `iter`, every affected cell is refitted and the message
names the component that changed.

Cells run in replication-major order: every row of replication 1, then
every row of replication 2, and so on. If a run is stopped early, the
finished cells already span the whole design.

## Seeds and simulated people

`seed` is a master seed. Each cell derives its own seed from it, its
row and its replication, so any single cell can be reproduced without
running the ones before it.

`subjects = "redraw"`, the default, draws new subject values in every
replication. `subjects = "fixed"` draws them once per row, in the first
replication, and reuses them in every later one, so the replications
become repeated simulations of the same people.

## When a cell fails

A cell whose fit throws an error is recorded with `status = "error"`,
the grid goes on, and a warning at the end lists the failed cells with
the first error message. A cell whose data cannot be generated stops the
grid instead, since that is an error in the design rather than in the
sampler.

## The result

The grid returns a single `bmmtools_recovery` object with population and
subject rows for every cell. The `condition` column names the grid row:

```{r}
recovery_mixture2p
```

`summary()` groups by condition. Population-level correlations are `NA`
here because every replication of a row uses the same population
values, so the truth has no spread within a condition; bias, RMSE and
coverage are still defined.

The attribute `cells` has one row per cell, with its seed, file, status,
runtime in seconds and convergence verdict:

```{r}
cells <- attr(recovery_mixture2p, "cells")
cells
```

To read the summary against the design, join the design columns from
`cells`:

```{r}
design <- unique(cells[c("condition", "n_subjects", "n_trials")])

subject_summary <- summary(recovery_mixture2p) |>
  dplyr::filter(level == "subject") |>
  dplyr::left_join(design, by = "condition")

subject_summary[c("term", "n_subjects", "n_trials", "n_converged",
                  "bias", "rmse", "coverage", "r", "r_low", "r_high")]
```

```{r, fig.height = 3.5}
library(ggplot2)
ggplot(
  subject_summary,
  aes(factor(n_trials), r, ymin = r_low, ymax = r_high,
      colour = factor(n_subjects))
) +
  geom_pointrange(position = position_dodge(width = 0.4)) +
  facet_wrap(~term) +
  labs(x = "Trials per subject", y = "Subject-level r (95% CI)",
       colour = "Subjects")
```

Because the object keeps its class through dplyr verbs, a single cell
can be plotted directly:

```{r}
recovery_mixture2p |>
  dplyr::filter(condition == "row-4", level == "subject") |>
  plot_recovery()
```

`elapsed` is the time a cell took in the run that returned the object,
in seconds, so the fitting time of the grid in minutes is:

```{r}
sum(cells$elapsed) / 60
```

After a resume, a cell whose fit was read from disk records the time it
took to read it, not the time it once took to fit.
