---
title: "Running a recovery study"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
options(width = 90)
# paths relative to the package root, whether this renders from the
# package root (pkgdown) or from vignettes/articles
from_root <- function(path) {
  if (dir.exists("vignettes")) path else file.path("..", "..", path)
}
# the code shown is read from the script that produced the results
shown <- function(label, file) {
  lines <- readLines(from_root(file))
  markers <- grep("^## ---- ", lines)
  start <- markers[lines[markers] == paste("## ----", label)]
  end <- c(markers[markers > start], length(lines) + 1L)[[1L]] - 1L
  code <- lines[(start + 1L):end]
  while (length(code) > 0L && !nzchar(trimws(code[[length(code)]]))) {
    code <- code[-length(code)]
  }
  code
}
library(bmmtools)
```

A recovery study repeats the loop of
[Validating a new bmm model](bmmtools.html), simulate, fit and score,
over a design: several numbers of subjects and trials, several
replications of each. `recovery_grid()` runs that loop and takes care of
what makes long simulations fragile: every cell is saved as soon as it
finishes, an interrupted run resumes from those files, and a short
preflight fit catches a broken model before hours are spent on it.

The example below is the grid behind `recovery_mixture2p`, the recovery
object that ships with the package. Its results are used live on this
page. The run took 8 minutes 58 seconds, compilation and preflight
included.

## The call

```{r grid, eval = FALSE, code = shown("grid", "data-raw/example-objects.R")}
```

Everything after `seed` is passed to `fit_cached()` and from there to
`bmm::bmm()`.

The call names `backend = "cmdstanr"`, and a grid should run on it.
cmdstanr compiles the model once and every cell with the same Stan code
reuses the executable, while rstan compiles again for every fit: in a
measurement on a small `mixture2p` grid, compiling took 21 to 23 seconds
per fit under rstan, 77 to 85 % of a cell's time, and once about 6
seconds under cmdstanr. Without a `backend` or the `brms.backend`
option, bmm picks cmdstanr when it is installed. A grid resumed in a new R session compiles once more,
because cmdstanr keeps executables in a temporary directory; setting
`options(cmdstanr_write_stan_file_dir = )` to a directory that persists
avoids that (see "Backend" in `?recovery_grid`).

## The design grid

`grid` is a data frame with one row per condition. It needs the columns
`n_subjects` and `n_trials`; any other column named after a parameter
sets that parameter's population value for the row, and a column
`sd_<parameter>` sets its between-subject SD. Both override `pars` and
`sds`, which act as the defaults for every row:

```{r, eval = FALSE}
grid <- expand.grid(
  n_subjects = c(20, 50),
  n_trials = c(30, 100),
  kappa = log(c(4, 16))
)
```

A design created with `SimDesign::createDesign()` is a data frame and
can be passed as it is. `model` can also be a function of one grid row
that returns a model, for designs where the model's arguments change
between rows; the link table must stay the same across rows, because
one table scores every cell.

## Smoke run and preflight

Before the full grid, run it with `smoke = TRUE`. This fits the first
two rows with two replications into `<dir>/smoke`, so a smoke run never
overwrites or gets mistaken for a full one.

By default, the grid also runs a **preflight**: before any cell, the
first cell is fitted once with one chain and 200 iterations into
`<dir>/preflight`. A compile error, an init failure or data the model
rejects stop the grid there, with nothing else run. The preflight is
skipped when the first cell already has a cached fit.

## Files on disk and resuming

Each cell writes two files as soon as it is done, named after its grid
row and replication:

- `cell-<row>-rep-<rep>-sim.rds` holds the simulated data and truth;
- `cell-<row>-rep-<rep>.rds` holds the fit, with its cache key in
  `cell-<row>-rep-<rep>.key`.

To resume after an interruption, run the same call again. A cell whose
simulation file exists reads it instead of simulating, and a cell whose
fit matches its key reuses the fit, so finished cells cost only the
time to read them. Rerunning the script that builds the example
objects, with all 20 grid fits on disk, took 23 seconds, including two
prior-only fits that were not cached yet. If an argument that enters the key
changes, say `iter`, every affected cell is refitted and the message
names the component that changed.

Cells run in replication-major order: every row of replication 1, then
every row of replication 2, and so on. If a run is stopped early, the
finished cells already span the whole design.

## Seeds and simulated people

`seed` is a master seed. Each cell derives its own seed from it, its
row and its replication, so any single cell can be reproduced without
running the ones before it.

`subjects = "redraw"`, the default, draws new subject values in every
replication. `subjects = "fixed"` draws them once per row, in the first
replication, and reuses them in every later one, so the replications
become repeated simulations of the same people.

## When a cell fails

A cell whose fit throws an error is recorded with `status = "error"`,
the grid goes on, and a warning at the end lists the failed cells with
the first error message. A cell whose data cannot be generated stops the
grid instead, since that is an error in the design rather than in the
sampler.

## Two estimators in one grid

`ml = TRUE` fits every cell a second time subject by subject with no
pooling, on the same simulated data, and scores both estimators against
the same truth at the subject level, so `summary()` returns a row per
estimator. `ml = list(...)` passes arguments to `fit_ml()`, and
`ml = list(method = "optim")` runs the second fit without a compiler.
The per-cell record is `attr(x, "ml_cells")`, and a cell whose ML fit
fails keeps its hierarchical rows. [Hierarchical estimation against
subject-wise maximum likelihood](hierarchical-vs-ml.html) works through
the comparison and what its columns mean.

## The result

The grid returns a single `bmmtools_recovery` object with population and
subject rows for every cell. The `condition` column names the grid row:

```{r}
recovery_mixture2p
```

`summary()` groups by condition. Population-level correlations are `NA`
here because every replication of a row uses the same population
values, so the truth has no spread within a condition; bias, RMSE and
coverage are still defined.

The attribute `cells` has one row per cell, with its seed, file, status,
runtime in seconds and convergence verdict:

```{r}
cells <- attr(recovery_mixture2p, "cells")
cells
```

To read the summary against the design, join the design columns from
`cells`:

```{r}
design <- unique(cells[c("condition", "n_subjects", "n_trials")])

subject_summary <- summary(recovery_mixture2p) |>
  dplyr::filter(level == "subject") |>
  dplyr::left_join(design, by = "condition")

subject_summary[c(
  "term", "n_subjects", "n_trials", "n_converged",
  "bias", "rmse", "coverage", "r", "r_low", "r_high"
)]
```

```{r, fig.height = 3.5}
library(ggplot2)
ggplot(
  subject_summary,
  aes(factor(n_trials), r,
    ymin = r_low, ymax = r_high,
    colour = factor(n_subjects)
  )
) +
  geom_pointrange(position = position_dodge(width = 0.4)) +
  facet_wrap(~term) +
  labs(
    x = "Trials per subject", y = "Subject-level r (95% CI)",
    colour = "Subjects"
  )
```

Because the object keeps its class through dplyr verbs, a single cell
can be plotted directly:

```{r}
recovery_mixture2p |>
  dplyr::filter(condition == "row-4", level == "subject") |>
  plot_recovery()
```

`elapsed` is the time a cell took in the run that returned the object,
in seconds, so the fitting time of the grid in minutes is:

```{r}
sum(cells$elapsed) / 60
```

After a resume, a cell whose fit was read from disk records the time it
took to read it, not the time it once took to fit.

## One grid, every estimand

A study usually needs more than one kind of number from the same data:
the population value of each parameter, an experimental effect, how much
people differ, whether those differences go together, and each
person's own values. `recovery_grid()` scores all of them from one fit
per cell. The result is then split back into one table per estimand
family, because a summary that pools subject values with population
effects answers no single question.

```{r combined-load, include = FALSE}
combined <- readRDS(from_root("vignettes/articles/assets/recovery-grid-combined.rds"))
combined_file <- "vignettes/articles/precompute/recovery-grid-combined.R"
families <- combined$families
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
pick <- function(table, condition, term, column) {
  table[[column]][table$condition == condition & table$term == term]
}
truth <- function(level, term) {
  rows <- combined$out[combined$out$level == level &
    combined$out$term == term, ]
  unique(rows$true_value)
}
cor_truth <- function(term) {
  unique(combined$correlations$true_value[combined$correlations$term == term])
}
cells_combined <- combined$cells
```

The example is `mixture2p` with two tasks. The second task lowers both
precision and the probability of recall. People vary around the task
values, and a person who is precise in one task tends to be precise in
the other:

```{r combined-model, eval = FALSE, code = shown("model", combined_file)}
```

`coding = "contrast"` fits an intercept and a task effect for each
parameter instead of one value per task, and `re_cor = "all"` lets
every subject deviation correlate with every other. Asking for the
`"effect"` and `"sd"` levels and for `correlations` adds those families
to the population and subject rows. Two cells, 40 and 100 subjects with
100 trials per task, ten replications each:

```{r combined-run, eval = FALSE, code = shown("run", combined_file)}
```

The data are generated per task. What the fit estimates is the
intercept (task 1, under the default treatment contrasts) and the
difference to task 2, so the grid transforms the generating values into
the truth for each of those terms. The two correlations of .7 between
tasks become a true SD of the `kappa` effect of
`r fmt(truth("sd", "kappa_task1"), 3)` and a true correlation of
`r fmt(cor_truth("kappa__kappa_task1"), 3)` between a person's `kappa`
intercept and their `kappa` effect; the correlations across parameters
are 0.

**Convergence.** The fits ran with bmm
`r combined$toolchain[["bmm"]]`, brms `r combined$toolchain[["brms"]]`
and CmdStan `r combined$toolchain[["cmdstan"]]`, 4 chains of 4000
iterations with 1000 of them warmup, and took
`r fmt(combined$minutes_fitting, 0)` minutes of sampling. All
`r sum(cells_combined$converged)` of `r nrow(cells_combined)` cells
converged: the largest Rhat of any fit was
`r fmt(max(cells_combined$max_rhat), 3)`, the smallest bulk ESS
`r fmt(min(cells_combined$min_ess_bulk), 0)` and the smallest tail ESS
`r fmt(min(cells_combined$min_ess_tail), 0)`. One fit had
`r sum(cells_combined$n_divergent)` divergent transition out of 12000.
An earlier run with 1000 iterations missed the convergence rule in most
cells; in its worst fit
the slowest variables were the correlations of the task effects, the
terms a design like this is least informative about.

### The families

`summary()` of the whole result has one row per cell, level and term.
Splitting it by `level` gives one table per family:

```{r combined-families, eval = FALSE, code = shown("families", combined_file)}
```

```{r combined-families-shown}
names(families)
```

Ten replications per cell set how finely the numbers below can be read:
a coverage over ten intervals moves in steps of .1, so .9 and 1 are
both consistent with a nominal 95 % interval, and a calibration slope
from ten replications shows a direction rather than a size (see
[Hierarchical estimation against subject-wise maximum
likelihood](hierarchical-vs-ml.html)).

**Effects** are scored on the link scale, the scale the contrast is
estimated on:

```{r combined-effects}
tibble::as_tibble(families$effect[c(
  "condition", "term", "bias", "rmse", "coverage", "coverage_50",
  "ci_width", "sign_recovery"
)])
```

The sign of both effects was recovered in every replication. The 95 %
intervals of the `kappa` effect narrow from
`r fmt(pick(families$effect, "row-1", "kappa_task1", "ci_width"))` with
40 subjects to `r fmt(pick(families$effect, "row-2", "kappa_task1", "ci_width"))`
with 100.

**Population values** are the intercepts, on the natural scale. Every
replication of a cell shares them, so there is no spread in the truth
and the correlation columns are `NA`, as in the grid above:

```{r combined-population}
tibble::as_tibble(families$population[c(
  "condition", "term", "bias", "rmse", "coverage", "coverage_50"
)])
```

**Between-subject SDs**, on the link scale:

```{r combined-sd}
tibble::as_tibble(families$sd[c(
  "condition", "term", "bias", "rmse", "coverage", "coverage_50",
  "ci_width"
)])
```

**Subject values** are summarised within each replication and then
combined, so here `r`, `ccc` and the calibration line are defined:

```{r combined-subject}
tibble::as_tibble(families$subject[c(
  "condition", "term", "rmse", "coverage", "r", "calibration_slope",
  "calibration_intercept"
)])
```

People are ordered better on their task-1 values than on how much they
change between tasks: with 100 subjects, `r` is
`r fmt(pick(families$subject, "row-2", "kappa", "r"))` for the `kappa`
intercept and `r fmt(pick(families$subject, "row-2", "kappa_task1", "r"))`
for the `kappa` effect, whose true SD is smaller. The calibration slope
of the `kappa` effect is above 1 in both cells
(`r fmt(pick(families$subject, "row-1", "kappa_task1", "calibration_slope"))`
and `r fmt(pick(families$subject, "row-2", "kappa_task1", "calibration_slope"))`),
so its subject estimates are shrunk more than the data warrant. With
ten replications that is the direction, not its size.

**Correlations** come as their own recovery object,
`attr(out, "correlations")`, with its own summary:

```{r combined-cors}
tibble::as_tibble(combined$correlations[c(
  "condition", "term", "true_value", "mean_estimate", "bias",
  "coverage"
)])
```

The correlations that are 0 in the truth stay near 0. The one between
a person's `kappa` intercept and `kappa` effect, truly
`r fmt(cor_truth("kappa__kappa_task1"))`, is estimated at
`r fmt(pick(combined$correlations, "row-1", "kappa__kappa_task1", "mean_estimate"))`
on average with 40 subjects and
`r fmt(pick(combined$correlations, "row-2", "kappa__kappa_task1", "mean_estimate"))`
with 100: pulled towards 0, less so with more people.
[Recovering between-subject correlations](correlation-recovery.html) looks at that
attenuation in detail.
